### 2015-09-10 manuscript.Rnw
initial commit

\documentclass[12pt,letterpaper]{article}

%%%%% packages provided by sys bio template
\usepackage{fixltx2e}
\usepackage{textcomp}
\usepackage{fullpage}
\usepackage{amsfonts}
\usepackage{verbatim}
%% \usepackage[english]{babel} %% I use polyglossia instead
%% not needed: \usepackage{pifont}
\usepackage{color}
\usepackage{setspace}
\usepackage{lscape}
\usepackage{indentfirst}
\usepackage[normalem]{ulem}
\usepackage{booktabs}
%\usepackage{nag}
\usepackage{natbib}
%\usepackage{bibtex}
\usepackage{float}
\usepackage{latexsym}
%\usepackage{hyperref}
\usepackage{url}
%\usepackage{html}
\usepackage{hyperref}
\usepackage{epsfig}
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{bm}
\usepackage{array}
%\usepackage{mhchem}
\usepackage{ifthen}
%% I am already using it: \usepackage{caption}
\usepackage{hyperref}
%\usepackage{xcolor}
\usepackage{amsthm}
\usepackage{amstext}
%%%%

\usepackage{natbib}
\usepackage{lineno}
\usepackage[T1]{fontenc}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage[font=sf]{caption}

% Latex special characters are rendered correctly with XeTeX
\usepackage{xltxtra}
\usepackage{xunicode}
\defaultfontfeatures{Mapping=tex-text}

% Words are cut where needed
\usepackage{polyglossia}
\setdefaultlanguage[variant=american]{english}

% Use fancy fonts
\usepackage{fontspec}
\setmainfont[Mapping=tex-text]{Crimson Text}
\setsansfont{SourceSansPro-Regular}

%%% syst bio options
\linespread{1.66}
% All text should be double-spaced
% with occasional exceptions for tables.
\raggedright
\setlength{\parindent}{0.5in}

\setcounter{secnumdepth}{0}

% Our sections are not numbered and our papers do not have
% Tables of Contents. We don't
% present a list of figures or list of tables, either.

% Any common font is fine.
% (A common sans-serif font should be used on figures, but figures should be
% separate from the LaTeX document.)

\pagestyle{empty}

\renewcommand{\section}[1]{%
\bigskip
\begin{center}
\begin{Large}
\normalfont\scshape #1
\medskip
\end{Large}
\end{center}}

\renewcommand{\subsection}[1]{%
\bigskip
\begin{center}
\begin{large}
\normalfont\itshape #1
\end{large}
\end{center}}

\renewcommand{\subsubsection}[1]{%
\vspace{2ex}
\noindent
\textit{#1.}---}

\renewcommand{\tableofcontents}{}

\bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib
%%% end syst bio options

\newcolumntype{b}{X}
\newcolumntype{s}{>{\hsize=.2\hsize}X}

% ----------------------------------------------------------------------------- %

\begin{document}

\begin{flushright}
Version dated: \today
\end{flushright}

\bigskip
\noindent AN R PACKAGE TO INTERACT WITH THE OPEN TREE OF LIFE

\bigskip
\medskip

\begin{center}
  \noindent{\Large \bf \code{rotl} an R package to interact with the Open Tree
    of Life Data}

\bigskip

\noindent{\normalsize \sc Fran\c{c}ois Michonneau}\\
\noindent {\small \it $^1$Florida Museum of Natural History & Whitney Laboratory
  for Marine Sciences, University of Florida, Gainesville,
  FL 32611-7800, USA}\\
\end{center}
\medskip

\noindent{\bf Corresponding author:} Fran\c{c}ois Michonneau, Division of
Invertebrate Zoology, Florida Museum of Natural History, Gainesville, FL
32611-7800, USA; E-mail: francois.michonneau@gmail.com\\

\vspace{1in}

\subsubsection{Abstract}

Abstract goes here.

\vspace{1.5in}

Phylogenetic information has become central to many facets of the life
sciences. However, the information is scattered across many publications, often
only available as image files and the variety of file formats used to store
phylogenetic information makes it difficult for researchers outside the field of
phylogenetics to access this information.

The Open Tree of Life (http://opentreeoflife.org) aims at assembling and
synthetizing our current understanding of the phylogenetic relationships across
all organisms on Earth \cite{Hinchliff2015}. This aims is achieved by combining
taxonomic information with published phylogenies. This is made possible by
mapping species found in published studies to species in the taxonomy, a
taxonomic lookup service is therefore needed. Presentation of the OTL:
synthesis, 2.3M tips, continuously updated.

This resource has many applications across the life sciences to explore
fundamental questions in evolution, biodiversity, conservation, and has great
potential for education and outreach purposes.

OTL also provides web interface to explore the open tree (and download
subtrees?)

Open Tree of Life provides an Application Programming Interface (API) to the
different services it offers. Here, we present \code{rotl} an R pacakge that
allows users to interact with the OTL's API to import phylogenetic trees and
taxonomic information in R where this information can be used to be analyzed
with the rich suites of packages available for phylogenetic analyses.

R has become a popular software and programming environment to analyze, explore
and visualize phylogenetic data because of the variety of packages available
\cite{}. ape, phytools, geiger, ggtree, phylobase, etc... (cite task view).

The OTL's API, and in turn \code{rotl}, provides services (1) to download a tree
from a set of species names, this can be useful to illustrate a manuscript, map
the evolution of a character in a taxonomic group, etc....; (2) to download a
tree published in a paper that might be used to build the synthetic tree, this
can be useful to reproduce or expand on an already published analysis; (3) to
download information about the taxonomy across all organisms considered by OTL
in a phylogenetic context.

Many other applications, getting trees for educational purpose (illustrating
course material, explore relationships among species), illustrate manuscript.

\section{General features}

The tree can be imported into R memory to interact with other packages and be
used in comparative analyses, or they can be written to files so they can
interact with other programs.

Interface with both the synthetic tree from Open Tree of Life and particular
trees from studies. Open Tree also features taxonomy lookup that can be used
with partial matching.

Limitations

No branch lengths for trees from the synthetic TOL, limit applications for
comparative data, but could be possible to use calibrations from time tree
databases.

Demonstration (1) getting relationships from a list of taxa

Demonstration (2) getting tree from a given study

* Which studies are available in OTL?
* How to look for them?

Demonstration (3) exploring the taxonomy

Demonstration (4) how does rotl fit into the R package ecosystem?

Conclusion

Resource as useful as the data that is in it. To make data available, it needs
to follow conventions so it can be incorporated into the OTL, i.e., submit tree
files (not images) to public repositories.

OTL provides an easy-to-use interface to submit studies.

Availability

`rotl` is available both from the CRAN repository
(https://cran.r-project.org/package=rotl) and from GitHub
(https://github.com/ropensci/rotl).

Acknowledgements

* OpenTree of Life hackathon
* Scott Chamberlain for code review.

\end{document}

### 2015-09-10 manuscript.Rnw
add LaTeX formatting for sections

@@
 
-Demonstration (1) getting relationships from a list of taxa
+\section{Demonstrations}
+\label{sec:demonstrations}
 
-Demonstration (2) getting tree from a given study
+\subsection{Getting relationships from a list of taxa}
+\label{sec:get-relationships}
+
+\subsection{Getting tree from a given study}
+\label{sec:get-tree-study}
 
@@
 
-Demonstration (3) exploring the taxonomy
+\subsection{Exploring the taxonomy}
+\label{sec:exploring-taxonomy}
 
-Demonstration (4) how does rotl fit into the R package ecosystem?
+\subsection{How does rotl fit into the R package ecosystem?}
+\label{sec:how-does-rotl-fit}
 
-Conclusion
+\section{Conclusion}
+\label{sec:conclusion}
 
@@
 
-Availability
+\section{Availability}
+\label{sec:availability}
 
-`rotl` is available both from the CRAN repository
+\code{rotl} is available both from the CRAN repository
 (https://cran.r-project.org/package=rotl) and from GitHub
@@
 
-Acknowledgements
+\section{Acknowledgments}
+\label{sec:acknowledgments}
 

### 2015-09-12 manuscript.Rnw
first editorial pass at introduction

@@
 \begin{center}
-  \noindent{\Large \bf \code{rotl} an R package to interact with the Open Tree
+  \noindent{\Large \bf \texttt{rotl} an R package to interact with the Open Tree
     of Life Data}
@@
 \noindent{\normalsize \sc Fran\c{c}ois Michonneau}\\
-\noindent {\small \it $^1$Florida Museum of Natural History & Whitney Laboratory
+\noindent {\small \it $^1$Florida Museum of Natural History \& Whitney Laboratory
   for Marine Sciences, University of Florida, Gainesville,
@@
 
-Phylogenetic information has become central to many facets of the life
-sciences. However, the information is scattered across many publications, often
-only available as image files and the variety of file formats used to store
-phylogenetic information makes it difficult for researchers outside the field of
-phylogenetics to access this information.
+Phylogenetic information data is being used in many disciplines of the
+life sciences, and phylogenetic comparative methods have seen their
+use diversified in recent years (community phylogenetics,
+...). However, phylogenetic information is scattered, often only
+available as image files within publications \cite{Mounce}, and the
+variety of file formats used to store phylogenetic information makes
+it difficult for researchers to access
 
-The Open Tree of Life (http://opentreeoflife.org) aims at assembling and
-synthetizing our current understanding of the phylogenetic relationships across
-all organisms on Earth \cite{Hinchliff2015}. This aims is achieved by combining
-taxonomic information with published phylogenies. This is made possible by
-mapping species found in published studies to species in the taxonomy, a
-taxonomic lookup service is therefore needed. Presentation of the OTL:
-synthesis, 2.3M tips, continuously updated.
+The Open Tree of Life (http://opentreeoflife.org) aims at assembling
+and synthesizing the current understanding of phylogenetic
+relationships across all organisms on Earth \cite{Hinchliff2015},
+while providing tools and services to allow users to access this
+information easily. OTL combines taxonomic information that serves as
+the backbone for the phylogenetic relationships, and published
+phylogenies to elucidate relationships among taxa.
 
-This resource has many applications across the life sciences to explore
-fundamental questions in evolution, biodiversity, conservation, and has great
-potential for education and outreach purposes.
+The current draft of the Open Tree of Life contains 2.3 million
+tips. Because of its comprehensive nature, researchers have access to
+phylogenetic information for species that have yet to be included in
+any phylogenetic tree as their position is inferred based on
+taxonomy. Beyond obvious applications across the life sciences to
+explore research questions in evolution, biodiversity, and
+conservation, the resources OTL provides could be used in education
+and outreach (e.g., illustrating course material, develop outreach
+activities to explore relationships among species).
 
-OTL also provides web interface to explore the open tree (and download
-subtrees?)
+The Open Tree of Life project provides a web interface to explore,
+query and download parts of the Open Tree
+(http://tree.opentreeoflife.org). Additionally, the project provides
+an Application Programming Interface (API) to the different services
+it offers allowing users to interact programmatically with the
+data. Here, we present \texttt{rotl} an R package that allows users to
+interact with the OTL's API to search for and import phylogenetic
+trees and taxonomic information. Having access to this data directly
+from within R allows researchers to take advantage of the multiple R
+packages that are available to visualize, analyze and manipulate
+phylogenetic trees (ape, phytools, geiger, ggtree, phylobase,
+etc... cite task view).
 
-Open Tree of Life provides an Application Programming Interface (API) to the
-different services it offers. Here, we present \code{rotl} an R pacakge that
-allows users to interact with the OTL's API to import phylogenetic trees and
-taxonomic information in R where this information can be used to be analyzed
-with the rich suites of packages available for phylogenetic analyses.
+\section{Presentation of the services provided by OTL}
 
-R has become a popular software and programming environment to analyze, explore
-and visualize phylogenetic data because of the variety of packages available
-\cite{}. ape, phytools, geiger, ggtree, phylobase, etc... (cite task view).
+Presentation of the OTL: synthesis, 2.3M tips, continuously
+updated. There are therefore 4 resources that the OTL provides: (1)
+the taxonomic information used as the backbone from the tree, (2) the
+trees and associated published studies that are used to build the
+synthetic tree, (3) a taxonomic name resolution service used to map
+species names to the Open Tree Taxonomy, (4) the synthetic tree. Also,
+Graph of Life endpoint but not of use for normal users and beyond the
+scope of this paper.
 
-The OTL's API, and in turn \code{rotl}, provides services (1) to download a tree
-from a set of species names, this can be useful to illustrate a manuscript, map
-the evolution of a character in a taxonomic group, etc....; (2) to download a
-tree published in a paper that might be used to build the synthetic tree, this
-can be useful to reproduce or expand on an already published analysis; (3) to
-download information about the taxonomy across all organisms considered by OTL
-in a phylogenetic context.
-
-Many other applications, getting trees for educational purpose (illustrating
-course material, explore relationships among species), illustrate manuscript.
-
-\section{General features}
-
-The tree can be imported into R memory to interact with other packages and be
-used in comparative analyses, or they can be written to files so they can
-interact with other programs.
+\texttt{rotl} provides access to these services so the data stored in
+OTL (including trees) can be imported into R (either in memory or by
+writing to files) to interact with other packages and be used in
+comparative analyses, or they can be written to files so they can used
+by other programs.
 
@@
 
+The OTL's API, and in turn \texttt{rotl}, provides services (1) to
+download a tree from a set of species names, this can be useful to
+illustrate a manuscript, map the evolution of a character in a
+taxonomic group, etc....; (2) to download a tree published in a paper
+that might be used to build the synthetic tree, this can be useful to
+reproduce or expand on an already published analysis; (3) to download
+information about the taxonomy across all organisms considered by OTL
+in a phylogenetic context.
+
+Tree versionning? Replicability of analyses
+
 Limitations
@@
 comparative data, but could be possible to use calibrations from time tree
-databases.
+databases. Or combined with partial chronograms and use \cite{phyndr}.
 
@@
 
-\code{rotl} is available both from the CRAN repository
+\texttt{rotl} is available both from the CRAN repository
 (https://cran.r-project.org/package=rotl) and from GitHub

### 2015-09-14 manuscript.Rnw
reworked intro, fix wrapping width

@@
 
-Phylogenetic information data is being used in many disciplines of the
-life sciences, and phylogenetic comparative methods have seen their
-use diversified in recent years (community phylogenetics,
-...). However, phylogenetic information is scattered, often only
-available as image files within publications \cite{Mounce}, and the
-variety of file formats used to store phylogenetic information makes
-it difficult for researchers to access
+Generating phylogenies has been facilitated by advances in sequencing and
+computing technologies enabling the investigation of evolutionary relationships
+among taxa. In parallel, new phylogenetic comparative methods have been
+developed and use these phylogenies to explore fundamental questions about the
+origin of biodiversity including the evolution of morphological and ecological
+traits, the spatio-temporal variation in speciation rates, or both
+\cite{Pennell2013}. The development of these methods has contributed to the need
+for a greater availability of high-quality phylogenetic information across
+life. However, phylogenetic information is scattered, often only available as
+image files within publications, and the variety of file formats used to store
+phylogenetic information makes it difficult for researchers to access and
+integrate into their own research (\cite{Stoltzfus2012}, but see
+\cite{Cranston2014} for suggestions of best practices).
 
-The Open Tree of Life (http://opentreeoflife.org) aims at assembling
-and synthesizing the current understanding of phylogenetic
-relationships across all organisms on Earth \cite{Hinchliff2015},
-while providing tools and services to allow users to access this
-information easily. OTL combines taxonomic information that serves as
-the backbone for the phylogenetic relationships, and published
-phylogenies to elucidate relationships among taxa.
+The Open Tree of Life (OTL) project aims at assembling and synthesizing our
+current understanding of phylogenetic relationships across all organisms on
+Earth \cite{Hinchliff2015}, while providing tools and services that facilitate
+access to this information. OTL combines taxonomic information that serves as
+the backbone for the phylogenetic relationships, and published phylogenies to
+elucidate relationships among taxa. This combination of information is used to
+form the synthetic tree. Studies can be contributed to the synthetic tree
+through the curator interface \url{https://tree.opentreeoflife.org/curator},
+allowing the tree to be updated continuously. The current draft of the Open Tree
+of Life contains 2.3 million tips. Beyond obvious applications across the life
+sciences to explore research questions in evolution, biodiversity, and
+conservation, the resources OTL provides could be used in education and outreach
+(e.g., illustrating course material, develop outreach activities to explore
+relationships among species).
 
-The current draft of the Open Tree of Life contains 2.3 million
-tips. Because of its comprehensive nature, researchers have access to
-phylogenetic information for species that have yet to be included in
-any phylogenetic tree as their position is inferred based on
-taxonomy. Beyond obvious applications across the life sciences to
-explore research questions in evolution, biodiversity, and
-conservation, the resources OTL provides could be used in education
-and outreach (e.g., illustrating course material, develop outreach
-activities to explore relationships among species).
+% Because of its comprehensive nature, researchers have access to phylogenetic
+% information for species that have yet to be included in any phylogenetic tree
+% as their position is inferred based on taxonomy.
 
-The Open Tree of Life project provides a web interface to explore,
-query and download parts of the Open Tree
-(http://tree.opentreeoflife.org). Additionally, the project provides
-an Application Programming Interface (API) to the different services
-it offers allowing users to interact programmatically with the
-data. Here, we present \texttt{rotl} an R package that allows users to
-interact with the OTL's API to search for and import phylogenetic
-trees and taxonomic information. Having access to this data directly
-from within R allows researchers to take advantage of the multiple R
-packages that are available to visualize, analyze and manipulate
-phylogenetic trees (ape, phytools, geiger, ggtree, phylobase,
-etc... cite task view).
+The Open Tree of Life project provides a web interface to explore, query and
+download parts of the synthetic Open Tree
+(\url{http://tree.opentreeoflife.org}). Additionally, the project provides an
+Application Programming Interface (API) to the different services it offers
+allowing users to work programmatically with the underlying data. Here, we
+present \texttt{rotl} an R package that allows users to interact with the OTL's
+API to search for and import phylogenetic trees and taxonomic
+information. Having access to this data directly from within R allows
+researchers to take advantage of the multiple R packages that are available to
+visualize, analyze and manipulate phylogenetic trees (ape, phytools, geiger,
+ggtree, phylobase, etc... cite task view).
 
@@
 
-Presentation of the OTL: synthesis, 2.3M tips, continuously
-updated. There are therefore 4 resources that the OTL provides: (1)
-the taxonomic information used as the backbone from the tree, (2) the
-trees and associated published studies that are used to build the
-synthetic tree, (3) a taxonomic name resolution service used to map
-species names to the Open Tree Taxonomy, (4) the synthetic tree. Also,
-Graph of Life endpoint but not of use for normal users and beyond the
-scope of this paper.
+Presentation of the OTL: synthesis, 2.3M tips, continuously updated. There are
+therefore 4 resources that the OTL provides: (1) the taxonomic information used
+as the backbone from the tree, (2) the trees and associated published studies
+that are used to build the synthetic tree, (3) a taxonomic name resolution
+service used to map species names to the Open Tree Taxonomy, (4) the synthetic
+tree. Also, Graph of Life endpoint but not of use for normal users and beyond
+the scope of this paper.
 
-\texttt{rotl} provides access to these services so the data stored in
-OTL (including trees) can be imported into R (either in memory or by
-writing to files) to interact with other packages and be used in
-comparative analyses, or they can be written to files so they can used
-by other programs.
+\texttt{rotl} provides access to these services so the data stored in OTL
+(including trees) can be imported into R (either in memory or by writing to
+files) to interact with other packages and be used in comparative analyses, or
+they can be written to files so they can used by other programs.
 
@@
 
-The OTL's API, and in turn \texttt{rotl}, provides services (1) to
-download a tree from a set of species names, this can be useful to
-illustrate a manuscript, map the evolution of a character in a
-taxonomic group, etc....; (2) to download a tree published in a paper
-that might be used to build the synthetic tree, this can be useful to
-reproduce or expand on an already published analysis; (3) to download
-information about the taxonomy across all organisms considered by OTL
+The OTL's API, and in turn \texttt{rotl}, provides services (1) to download a
+tree from a set of species names, this can be useful to illustrate a manuscript,
+map the evolution of a character in a taxonomic group, etc....; (2) to download
+a tree published in a paper that might be used to build the synthetic tree, this
+can be useful to reproduce or expand on an already published analysis; (3) to
+download information about the taxonomy across all organisms considered by OTL
 in a phylogenetic context.

### 2015-09-15 manuscript.Rnw
edits to presentation of OTL's APIs

@@
 
-Generating phylogenies has been facilitated by advances in sequencing and
-computing technologies enabling the investigation of evolutionary relationships
-among taxa. In parallel, new phylogenetic comparative methods have been
-developed and use these phylogenies to explore fundamental questions about the
-origin of biodiversity including the evolution of morphological and ecological
-traits, the spatio-temporal variation in speciation rates, or both
-\cite{Pennell2013}. The development of these methods has contributed to the need
-for a greater availability of high-quality phylogenetic information across
-life. However, phylogenetic information is scattered, often only available as
-image files within publications, and the variety of file formats used to store
-phylogenetic information makes it difficult for researchers to access and
-integrate into their own research (\cite{Stoltzfus2012}, but see
-\cite{Cranston2014} for suggestions of best practices).
+Generating phylogenies has been greatly facilitated by advances in sequencing
+and computing technologies enabling the investigation of evolutionary
+relationships among taxa. In parallel, new phylogenetic comparative methods have
+been developed and make use of these phylogenies to explore fundamental
+questions about the origin of biodiversity including the evolution of
+morphological and ecological traits, the spatio-temporal variation in speciation
+rates, or both \cite{Pennell2013}. The development of these methods has
+contributed to the need for a greater availability of high-quality phylogenetic
+information across life. However, phylogenetic information is scattered, often
+only available as image files within publications, and the variety of file
+formats used to store phylogenetic data makes it difficult for researchers to
+access, synthesize, and integrate into their own research (\cite{Stoltzfus2012},
+but see \cite{Cranston2014} for suggestions of best practices).
 
@@
 through the curator interface \url{https://tree.opentreeoflife.org/curator},
-allowing the tree to be updated continuously. The current draft of the Open Tree
-of Life contains 2.3 million tips. Beyond obvious applications across the life
-sciences to explore research questions in evolution, biodiversity, and
-conservation, the resources OTL provides could be used in education and outreach
-(e.g., illustrating course material, develop outreach activities to explore
-relationships among species).
+allowing the tree to be updated as relationships are elucidated or
+reevaluated. The current draft of the Open Tree of Life contains 2.3 million
+tips. Beyond obvious applications across the life sciences to explore research
+questions in evolution, biodiversity, and conservation, the resources OTL
+provides could be used in education and outreach (e.g., illustrating course
+material, develop outreach activities to explore relationships among species).
 
@@
 allowing users to work programmatically with the underlying data. Here, we
-present \texttt{rotl} an R package that allows users to interact with the OTL's
-API to search for and import phylogenetic trees and taxonomic
-information. Having access to this data directly from within R allows
-researchers to take advantage of the multiple R packages that are available to
-visualize, analyze and manipulate phylogenetic trees (ape, phytools, geiger,
-ggtree, phylobase, etc... cite task view).
+present \texttt{rotl} an R package that allows users to interact with OTL's API
+to search for, and import phylogenetic trees and taxonomic information into
+R. Having access to this data directly in R allows researchers to take advantage
+of the multiple R packages that are available to visualize, analyze and
+manipulate phylogenetic and comparative data (ape, phytools, geiger, ggtree,
+phylobase, RNeXML, etc... cite task view).
 
-\section{Presentation of the services provided by OTL}
+\section{Services provided by OTL}
 
-Presentation of the OTL: synthesis, 2.3M tips, continuously updated. There are
-therefore 4 resources that the OTL provides: (1) the taxonomic information used
-as the backbone from the tree, (2) the trees and associated published studies
-that are used to build the synthetic tree, (3) a taxonomic name resolution
-service used to map species names to the Open Tree Taxonomy, (4) the synthetic
-tree. Also, Graph of Life endpoint but not of use for normal users and beyond
-the scope of this paper.
+The OTL project provides four resources that serve data to users through the API
+(the API endpoints):
 
-\texttt{rotl} provides access to these services so the data stored in OTL
-(including trees) can be imported into R (either in memory or by writing to
-files) to interact with other packages and be used in comparative analyses, or
-they can be written to files so they can used by other programs.
+\begin{enumerate}
+\item The taxonomic information used as the backbone from the tree, the Open
+  Tree Taxonomy;
+\item The studies and their associated trees some of which are chosen by
+  curators to assemble the synthetic tree;
+\item A taxonomic name resolution service (TNRS) used to map taxon names to
+  the Open Tree Taxonomy;
+\item The synthetic tree itself (the Open Tree).
+\end{enumerate}
 
-Interface with both the synthetic tree from Open Tree of Life and particular
-trees from studies. Open Tree also features taxonomy lookup that can be used
-with partial matching.
+Additionally, the project provides access to other data, the Graph of Life, that
+stores information about the underlying tree alignment graphs \cite{Smith2013}
+used to synthesize the various sources of information from which the Open Tree
+is assembled from. Data from this API endpoint is not intended for normal users
+and its use will not be demonstrated in this paper but \texttt{rotl} has
+functions to retrieve data it provides.
 
-The OTL's API, and in turn \texttt{rotl}, provides services (1) to download a
-tree from a set of species names, this can be useful to illustrate a manuscript,
-map the evolution of a character in a taxonomic group, etc....; (2) to download
-a tree published in a paper that might be used to build the synthetic tree, this
-can be useful to reproduce or expand on an already published analysis; (3) to
-download information about the taxonomy across all organisms considered by OTL
-in a phylogenetic context.
+\texttt{rotl} gives users access to all endpoints provided by version 2 of the
+OTL's API. Phylogenetic trees served by the API can be imported directly into
+R's memory and are represented using the \texttt{ape} \cite{Paradis2004} tree
+structure (objects of class \texttt{phylo}) or can be written to files in the
+NEXUS, Newick or NeXML file formats. This allows researchers to use these trees
+either directly with other R packages, or to be imported in other programs that
+can read in phylogenetic tree files.
 
-Tree versionning? Replicability of analyses
+The synthetic tree currently does not have any branch lengths associated with
+it, therefore parametric comparative methods cannot be used on the trees
+returned by the API. However, resources and methods are being developed to add
+branch lengths information to these topological trees (e.g., \cite{Ksepka2015},
+\cite{Pennell2015}). Even without branch lengths, these trees could be useful to
+illustrate relationships among species, or to map a trait on a phylogeny.
 
-Limitations
+\texttt{rotl} can also be used to download trees accompanying studies, and that
+are used to build the synthetic tree. All the trees from a study are available
+even if they do not contribute to the synthetic tree. These trees provide a
+useful resource to reproduce or expand on an already published analysis.
+%% any examples available?
 
-No branch lengths for trees from the synthetic TOL, limit applications for
-comparative data, but could be possible to use calibrations from time tree
-databases. Or combined with partial chronograms and use \cite{phyndr}.
+% Tree versionning? Replicability of analyses
+
+\section{Technical information about \texttt{rotl}}
+
+Use NCL/\texttt{rncl} to parse NEXUS/Newick files.
+
+Taxonomic information, singletons, but NCL can parse it.
 

### 2015-09-24 manuscript.Rnw
get line numbers to work
small edits to intro

@@
 
+% line numbers
+\usepackage{lineno}
+\modulolinenumbers[5]
+\linenumbers
+
 \usepackage{natbib}
-\usepackage{lineno}
 \usepackage[T1]{fontenc}
@@
 through the curator interface \url{https://tree.opentreeoflife.org/curator},
-allowing the tree to be updated as relationships are elucidated or
-reevaluated. The current draft of the Open Tree of Life contains 2.3 million
-tips. Beyond obvious applications across the life sciences to explore research
-questions in evolution, biodiversity, and conservation, the resources OTL
-provides could be used in education and outreach (e.g., illustrating course
-material, develop outreach activities to explore relationships among species).
+allowing the synthetic tree to be continuously updated as relationships are
+elucidated or reevaluated. The current draft of the Open Tree of Life contains
+2.3 million tips. Beyond obvious applications across the life sciences to
+explore research questions in evolution, biodiversity, and conservation, the
+resources OTL provides could be used in education and outreach (e.g.,
+illustrating course material, develop outreach activities to explore
+relationships among species).
 
@@
 \texttt{rotl} gives users access to all endpoints provided by version 2 of the
-OTL's API. Phylogenetic trees served by the API can be imported directly into
+OTL's APIs. Phylogenetic trees served by the API can be imported directly into
 R's memory and are represented using the \texttt{ape} \cite{Paradis2004} tree
-structure (objects of class \texttt{phylo}) or can be written to files in the
-NEXUS, Newick or NeXML file formats. This allows researchers to use these trees
-either directly with other R packages, or to be imported in other programs that
-can read in phylogenetic tree files.
+structure (objects of class \texttt{phylo}), or can be written to files in the
+Newick, NEXUS \citep{Maddison1997} or NeXML \citep{Vos2012} file formats. This
+allows researchers to use these trees either directly with other R packages, or
+to be imported in other programs that can read in phylogenetic tree files.
 
@@
 returned by the API. However, resources and methods are being developed to add
-branch lengths information to these topological trees (e.g., \cite{Ksepka2015},
-\cite{Pennell2015}). Even without branch lengths, these trees could be useful to
+branch lengths information to these topological trees (e.g., \cite{Ksepka2015})
+or use topological trees to identify phylogenetically equivalent species to
+increase overlap between chronograms and species trait data
+\cite{Pennell2015}. Without branch lengths, these trees remain useful to
 illustrate relationships among species, or to map a trait on a phylogeny.
 
-\texttt{rotl} can also be used to download trees accompanying studies, and that
-are used to build the synthetic tree. All the trees from a study are available
-even if they do not contribute to the synthetic tree. These trees provide a
-useful resource to reproduce or expand on an already published analysis.
+\texttt{rotl} can also be used to download trees accompanying studies that have
+been submitted through the curator interface, and identify the trees that are
+being used to assemble the synthetic tree. These trees provide a useful resource
+to reproduce or expand on an already published analysis.
+
 %% any examples available?

### 2015-09-24 manuscript.Rnw
attempt at technical details section

@@
 
-Use NCL/\texttt{rncl} to parse NEXUS/Newick files.
-
-Taxonomic information, singletons, but NCL can parse it.
+Phylogenetic information retrieved from OTL is converted into
+\texttt{ape::phylo} objects by \texttt{rotl} using the NEXUS Class Library
+\citep[NCL,][]{Lewis2003} as implemented in the \texttt{rncl} package
+\url{https://cran.r-project.org/package=rncl}. Using NCL provides efficient
+parsing of large trees that may contain singletons nodes labeled with taxonomic
+information.
 
@@
 
+Requesting a topology from a list of taxa takes place in two steps: the taxa
+need to be matched to their identifiers in the Open Tree Taxonomy before
+requesting the relationships among them. The first step is performed using the
+TNRS.
+
 \subsection{Getting tree from a given study}
@@
 * How to look for them?
+
+<<>>=
+tt <- studies_find_studies(property = "ot:focalCladeOTTTaxonName", "mammalia")
+@
 

### 2015-09-29 manuscript.Rnw
starting to work on the examples

--- 2015-09-24 manuscript.Rnw
@@
 Generating phylogenies has been greatly facilitated by advances in sequencing
-and computing technologies enabling the investigation of evolutionary
-relationships among taxa. In parallel, new phylogenetic comparative methods have
-been developed and make use of these phylogenies to explore fundamental
+and computing technologies. In parallel, new phylogenetic comparative methods
+have been developed and make use of these phylogenies to explore fundamental
 questions about the origin of biodiversity including the evolution of
@@
 formats used to store phylogenetic data makes it difficult for researchers to
-access, synthesize, and integrate into their own research (\cite{Stoltzfus2012},
-but see \cite{Cranston2014} for suggestions of best practices).
+access, synthesize, and integrate this information into their own research
+(\cite{Stoltzfus2012}, but see \cite{Cranston2014} for suggestions of best
+practices).
 
@@
 explore research questions in evolution, biodiversity, and conservation, the
-resources OTL provides could be used in education and outreach (e.g.,
+resources OTL provides could be used for education and outreach (e.g.,
 illustrating course material, develop outreach activities to explore
@@
 (\url{http://tree.opentreeoflife.org}). Additionally, the project provides an
-Application Programming Interface (API) to the different services it offers
-allowing users to work programmatically with the underlying data. Here, we
-present \texttt{rotl} an R package that allows users to interact with OTL's API
-to search for, and import phylogenetic trees and taxonomic information into
-R. Having access to this data directly in R allows researchers to take advantage
-of the multiple R packages that are available to visualize, analyze and
-manipulate phylogenetic and comparative data (ape, phytools, geiger, ggtree,
-phylobase, RNeXML, etc... cite task view).
+Application Programming Interface (API) to the different services to work
+programmatically with the underlying data. Here, we present \texttt{rotl} an R
+package to interact with OTL's API using R syntax to search for, and import
+phylogenetic trees and taxonomic information into R. Having access to this data
+directly in R allows researchers to take advantage of the multiple packages that
+are available to visualize, analyze and manipulate phylogenetic and comparative
+data (ape, phytools, geiger, ggtree, phylobase, RNeXML, etc... cite task view
+\url{https://cran.r-project.org/web/views/Phylogenetics.html}).
 
@@
 
-The OTL project provides four resources that serve data to users through the API
-(the API endpoints):
+The OTL project provides four resources that serve data to users through the
+API:
 
 \begin{enumerate}
-\item The taxonomic information used as the backbone from the tree, the Open
-  Tree Taxonomy;
-\item The studies and their associated trees some of which are chosen by
+\item The \emph{taxonomy} used as the backbone from the tree, the Open Tree
+  Taxonomy;
+\item The \emph{studies} and their associated trees some of which are chosen by
   curators to assemble the synthetic tree;
-\item A taxonomic name resolution service (TNRS) used to map taxon names to
-  the Open Tree Taxonomy;
-\item The synthetic tree itself (the Open Tree).
+\item A \emph{taxonomic name resolution service} (TNRS) used to match taxon
+  names to the Open Tree Taxonomy identifiers;
+\item The \emph{synthetic tree} itself (the Open Tree).
 \end{enumerate}
@@
 The synthetic tree currently does not have any branch lengths associated with
-it, therefore parametric comparative methods cannot be used on the trees
-returned by the API. However, resources and methods are being developed to add
-branch lengths information to these topological trees (e.g., \cite{Ksepka2015})
-or use topological trees to identify phylogenetically equivalent species to
-increase overlap between chronograms and species trait data
-\cite{Pennell2015}. Without branch lengths, these trees remain useful to
-illustrate relationships among species, or to map a trait on a phylogeny.
-
-\texttt{rotl} can also be used to download trees accompanying studies that have
-been submitted through the curator interface, and identify the trees that are
-being used to assemble the synthetic tree. These trees provide a useful resource
-to reproduce or expand on an already published analysis.
+it, therefore parametric comparative methods cannot be used directly on the
+trees returned by the API. However, resources and methods are being developed to
+add branch lengths information to these topological trees (e.g.,
+\cite{Ksepka2015}) or use topological trees to identify phylogenetically
+equivalent species to increase overlap between chronograms and species trait
+data \cite{Pennell2015}. Without branch lengths, these trees remain useful to
+illustrate relationships among species, or to map a trait on a phylogeny for
+instance.
 
@@
 
-Use NCL/\texttt{rncl} to parse NEXUS/Newick files.
-
-Taxonomic information, singletons, but NCL can parse it.
+Phylogenetic information retrieved from OTL is converted into
+\texttt{ape::phylo} objects by \texttt{rotl} using the NEXUS Class Library
+\citep[NCL,][]{Lewis2003} as implemented in the \texttt{rncl} package
+\url{https://cran.r-project.org/package=rncl}. Using NCL provides robust and
+efficient parsing of large trees that may contain singletons nodes labeled with
+taxonomic information.
 
@@
 
-\subsection{Getting tree from a given study}
+To get the relationships among a set of taxa from the Open Tree, the taxa first
+need to be matched against the Open Taxonomy using the TNRS. This step retrieves
+the taxa identifiers that will be used to extract the relationships for the set
+of requested taxa.
+
+To illustrate how to obtain relationships from a set of taxa, we will rely on
+OTL to draw a phylogenetic tree for a set of model organisms.
+
+<<init, echo=FALSE>>=
+library(rotl)
+@
+
+First, we use the function \texttt{tnrs\_match\_names} to match the taxon names to
+their Open Taxonomy identifiers.
+
+<<get_taxa, cache=TRUE>>=
+taxa <- tnrs_match_names(names = c("Escherichia coli",
+                                   "Chlamydomonas reinhardtii",
+                                   "Drosophila melanogaster",
+                                   "Arabidopsis thaliana",
+                                   "Rattus norvegicus",
+                                   "Mus musculus",
+                                   "Cavia porcellus",
+                                   "Xenopus laevis",
+                                   "Saccharomyces cervisae",
+                                   "Danio rerio"))
+@
+
+The function \texttt{tnrs\_match\_names} returns a data frame that lists the Open
+Tree identifiers as well as other information to help users to ensure that the
+taxa matched are the correct ones. Here, there is no ambiguity in the taxa
+matched, however, as the Open Tree Taxonomy includes taxa from bacteria, plants,
+and animals that are regulated by different nomenclatural codes (ICNP, ICN and
+ICZN respectively), OTL and \texttt{rotl} provide tools to deal with potential
+hemihomonyms. For instance, the argument \texttt{context\_name} can be used to
+limit potential matches to a taxonomic group. The function
+\texttt{tnrs\_contexts} provides a list of possible taxonomic contexts. When this
+strategy cannot be used in case the tree encompasses multiple domains, the
+functions \texttt{inspect} and \texttt{update} list alternative matches for a
+taxon name, and replace it in the results, respectively. An example of this
+approach is provided in the vignette ``How to use rotl?'' that accompanies the
+package.
+
+Now that the taxon names are matched to the Open Tree identifiers, we can pass
+them to the function \texttt{tol\_induced\_subtree} to retrieve the
+relationships among these taxa. In turn, the tree can be plotted directly as it
+is returned as an \texttt{ape::phylo} object (Figure~\ref{fig:plot_taxa}).
+
+<<plot_taxa, cache=TRUE, fig.cap="The phylogenetic tree returned by OTL for the list of model species used as an example.">>=
+tree <- tol_induced_subtree(ott_ids = taxa[["ott_id"]])
+plot(tree)
+@
+
+\subsection{Getting trees from studies}
 \label{sec:get-tree-study}
 
-* Which studies are available in OTL?
-* How to look for them?
+\texttt{rotl} can also be used to retrieve trees accompanying studies that have
+been submitted through the curator interface, and also identify the trees that
+are being used to assemble the synthetic tree. These trees provide a useful
+resource to reproduce or expand on an already published analysis.
+
+Criteria that can be used to search for studies or their associated trees are
+available through the output of the function
+\texttt{studies\_properties}. Typically, users will want to search for studies
+or trees based on a taxon but other criteria such as the title of the
+publication. Here we demonstrate how to look for and retrieve a tree for study
+focusing on cats, Felidae (Figure~\ref{fig:plot_cats}).
+
+<<plot_cats, fig.cap="Tree from \\cite{} obtained from OTL">>=
+cat_studies <- studies_find_studies(property = "ot:focalCladeOTTTaxonName",
+                                    value = "felidae")
+cat_studies
+cat_tree <- get_study_tree(study_id = cat_studies[["study_ids"]][1],
+                           tree_id = list_trees(cat_studies)[[1]][1])
+plot(cat_tree)
+@
 
@@
 
+\bibliographystyle{sysbio}
+
+\bibliography{rotl-manuscript}
+
 \end{document}

### 2015-09-29 manuscript.Rnw
finish studies example, fix citations

@@
 morphological and ecological traits, the spatio-temporal variation in speciation
-rates, or both \cite{Pennell2013}. The development of these methods has
+rates, or both \citep{Pennell2013}. The development of these methods has
 contributed to the need for a greater availability of high-quality phylogenetic
@@
 access, synthesize, and integrate this information into their own research
-(\cite{Stoltzfus2012}, but see \cite{Cranston2014} for suggestions of best
+(\citealt{Stoltzfus2012}, but see \citealt{Cranston2014} for suggestions of best
 practices).
@@
 current understanding of phylogenetic relationships across all organisms on
-Earth \cite{Hinchliff2015}, while providing tools and services that facilitate
+Earth \citep{Hinchliff2015}, while providing tools and services that facilitate
 access to this information. OTL combines taxonomic information that serves as
@@
 OTL's APIs. Phylogenetic trees served by the API can be imported directly into
-R's memory and are represented using the \texttt{ape} \cite{Paradis2004} tree
+R's memory and are represented using the \texttt{ape} \citep{Paradis2004} tree
 structure (objects of class \texttt{phylo}), or can be written to files in the
@@
 trees returned by the API. However, resources and methods are being developed to
-add branch lengths information to these topological trees (e.g.,
-\cite{Ksepka2015}) or use topological trees to identify phylogenetically
-equivalent species to increase overlap between chronograms and species trait
-data \cite{Pennell2015}. Without branch lengths, these trees remain useful to
-illustrate relationships among species, or to map a trait on a phylogeny for
-instance.
+add branch lengths information to these topological trees
+\citep[e.g.,][]{Ksepka2015} or use topological trees to identify
+phylogenetically equivalent species to increase overlap between chronograms and
+species trait data \citep{Pennell2015}. Without branch lengths, these trees
+remain useful to illustrate relationships among species, or to map a trait on a
+phylogeny for instance.
 
@@
 \texttt{rotl} can also be used to retrieve trees accompanying studies that have
-been submitted through the curator interface, and also identify the trees that
-are being used to assemble the synthetic tree. These trees provide a useful
-resource to reproduce or expand on an already published analysis.
+been submitted through the curator interface, and identify the trees that
+contribute to the synthetic tree. These trees provide a useful resource to
+reproduce or expand on an already published analysis.
 
@@
 \texttt{studies\_properties}. Typically, users will want to search for studies
-or trees based on a taxon but other criteria such as the title of the
-publication. Here we demonstrate how to look for and retrieve a tree for study
-focusing on cats, Felidae (Figure~\ref{fig:plot_cats}).
+or trees based on taxon names (or their Open Taxonomy identifiers), but other
+criteria such as the title of the publication can be used. Here we demonstrate
+how to look for and retrieve a tree for study focusing on the family Felidae
+(Figure~\ref{fig:plot_cats}).
 
-<<plot_cats, fig.cap="Tree from \\cite{} obtained from OTL">>=
+<<get_cats, fig.cap="Tree from \\cite{} obtained from OTL">>=
 cat_studies <- studies_find_studies(property = "ot:focalCladeOTTTaxonName",
@@
 cat_studies
-cat_tree <- get_study_tree(study_id = cat_studies[["study_ids"]][1],
-                           tree_id = list_trees(cat_studies)[[1]][1])
-plot(cat_tree)
 @
 
-\subsection{Exploring the taxonomy}
-\label{sec:exploring-taxonomy}
+Currently only one study focused on this family is available from OTL, and a
+single tree is associated with it. We can then retrieve the identifiers and pass
+them to the function \texttt{get\_study\_tree} to have the tree in memory:
 
-\subsection{How does rotl fit into the R package ecosystem?}
+<<get_cat_tree, results='hide'>>=
+cat_tree <- get_study_tree(study_id = cat_studies[["study_ids"]][1],
+                           tree_id = cat_studies[["tree_ids"]][1])
+cat_tree
+@
+
+<<print_cat_tree, echo=FALSE>>=
+print(cat_tree, printlen = 4)
+@
+
+When more than one tree is available for a given study, the function
+\texttt{list\_trees} returns a list to retrieve the tree identifiers for each
+study easily. Alternatively, the function \texttt{get\_study} returns all the
+trees associated with a particular study. Additional details about the study can
+be obtained using the function \texttt{get\_study\_meta}.
+
+\subsection{How does \texttt{rotl} fit into the R package ecosystem?}
 \label{sec:how-does-rotl-fit}

### 2015-10-01 manuscript.Rnw
add example on how to use rotl with rgbif

@@
 
+In recent years, R has become part of the toolbox of many researchers in
+evolutionary biology. As many methods for comparative analyses are implemented,
+and because it is a relatively easy to use programming language, R greatly
+facilitates the analysis of large datasets, and allows researchers to combine
+methods in novel ways. Additionally, as more data is made available online and
+accessible using APIs, many R packages have been developed to take advantage of
+these datasets, especially spearheaded by the organization rOpenSci
+\url{https://ropensci.org}. \texttt{rotl} contributes to these initiatives and
+data from OTL can be combined with other sources easily. For instance in
+Figure~\ref{cat_map}, we show how we can obtain a map of the occurrences for
+some of the cat species (genus \textit{Lynx}) included in the phylogeny above
+using records for these species found in GBIF (code is available in
+supplementary material).
+
+<<cat_map, echo=FALSE, results='hide', message=FALSE, cache=TRUE, fig.cap='GBIF records for the species in \\textit{Lynx} included in the phylogeny associated with the study by \\citealt{Johnson2006}'>>=
+
+suppressPackageStartupMessages(library(rgbif))
+suppressPackageStartupMessages(library(ggplot2))
+suppressPackageStartupMessages(library(maps))
+
+cat_species <- cat_tree$tip.label
+lynx_species <- grep("^Lynx", cat_tree$tip.label, value = TRUE)
+
+gbif_keys <- sapply(lynx_species,
+                    function(x) name_backbone(name = x)$speciesKey,
+                    USE.NAMES = FALSE)
+
+lynx_loc <- occ_search(taxonKey = gbif_keys, limit = 500,
+                       return = "data", fields = "minimal",
+                       hasCoordinate = TRUE)
+
+lynx_loc <- do.call("rbind", lynx_loc)
+lynx_loc[["decimalLatitude"]] <- as.numeric(lynx_loc[["decimalLatitude"]])
+lynx_loc[["decimalLongitude"]] <- as.numeric(lynx_loc[["decimalLongitude"]])
+lynx_loc <- lynx_loc[complete.cases(lynx_loc), ]
+
+names(lynx_loc)[1] <- "Species"
+
+world <- map_data("world")
+
+ggplot(lynx_loc) +
+    annotation_map(world, fill="gray40", color="gray40") +
+    geom_point(aes(y = decimalLatitude, x = decimalLongitude, color = Species),
+               alpha = 0.3, size = 2) +
+    coord_map(projection = "mercator", orientation = c(90, 0, 0)) +
+    xlab("Longitude") + ylab("Latitude")
+
+@
+
 \section{Conclusion}

### 2015-10-01 manuscript.Rnw
add paragraph about fuzzy matching + synonyms

@@
 package.
+By default, approximate matching is enabled when attempting to match taxonomic
+names to their Open Tree Taxonomy identifiers. Additionally, taxonomic synonyms
+are included in the Open Tree Taxonomy allowing researchers to match correct
+identifiers for taxon names that might include misspellings or synonyms. These
+features will facilitate the tedious data cleaning process often needed when
+matching taxon names. In the example provided, both \textit{Escherichia coli}
+and \textit{Saccharomyces cerevisiae} are misspelled but OTL's TNRS finds the
+correct match for these taxa.
 

### 2015-10-01 manuscript.Rnw
various edits

@@
 <<get_taxa, cache=TRUE>>=
-taxa <- tnrs_match_names(names = c("Escherichia coli",
+taxa <- tnrs_match_names(names = c("Escherichia colli",
                                    "Chlamydomonas reinhardtii",
@@
 
-The function \texttt{tnrs\_match\_names} returns a data frame that lists the Open
-Tree identifiers as well as other information to help users to ensure that the
-taxa matched are the correct ones. Here, there is no ambiguity in the taxa
+The function \texttt{tnrs\_match\_names} returns a data frame that lists the
+Open Tree identifiers as well as other information to help users to ensure that
+the taxa matched are the correct ones. Here, there is no ambiguity in the taxa
 matched, however, as the Open Tree Taxonomy includes taxa from bacteria, plants,
@@
 ICZN respectively), OTL and \texttt{rotl} provide tools to deal with potential
-hemihomonyms. For instance, the argument \texttt{context\_name} can be used to
-limit potential matches to a taxonomic group. The function
-\texttt{tnrs\_contexts} provides a list of possible taxonomic contexts. When this
-strategy cannot be used in case the tree encompasses multiple domains, the
-functions \texttt{inspect} and \texttt{update} list alternative matches for a
-taxon name, and replace it in the results, respectively. An example of this
-approach is provided in the vignette ``How to use rotl?'' that accompanies the
-package.
+hemihomonyms. The argument \texttt{context\_name} can be used to limit potential
+matches to a taxonomic group (see the function \texttt{tnrs_contexts} for a list
+of possible options).  When this strategy cannot be used in case the tree
+encompasses multiple domains (as in the present example), the function
+\texttt{inspect} lists alternative matches for a taxon name, and \texttt{update}
+replaces it in the results. An example of this approach is provided in the
+vignette ``How to use rotl?'' that accompanies the package.
+
 By default, approximate matching is enabled when attempting to match taxonomic
@@
 Currently only one study focused on this family is available from OTL, and a
-single tree is associated with it. We can then retrieve the identifiers and pass
-them to the function \texttt{get\_study\_tree} to have the tree in memory:
+single tree is associated with it. We can then retrieve the study and tree
+identifiers, and pass them to the function \texttt{get\_study\_tree} to have the
+tree in memory:
 
@@
 When more than one tree is available for a given study, the function
-\texttt{list\_trees} returns a list to retrieve the tree identifiers for each
+\texttt{list\_trees} returns a list containing the tree identifiers for each
 study easily. Alternatively, the function \texttt{get\_study} returns all the
-trees associated with a particular study. Additional details about the study can
-be obtained using the function \texttt{get\_study\_meta}.
+trees (by default as \texttt{ape::phylo} objects) associated with a particular
+study. Additional details about the study (i.e., its metadata) can be obtained
+using the function \texttt{get\_study\_meta}.
 

### 2015-10-04 manuscript.Rnw
tweaks to makefile and MS for easier building

@@
 hemihomonyms. The argument \texttt{context\_name} can be used to limit potential
-matches to a taxonomic group (see the function \texttt{tnrs_contexts} for a list
+matches to a taxonomic group (see the function \texttt{tnrs\_contexts} for a list
 of possible options).  When this strategy cannot be used in case the tree

### 2015-10-09 manuscript.Rnw
add Joseph and David as authors

--- 2015-10-01 manuscript.Rnw
@@
 
-\noindent{\normalsize \sc Fran\c{c}ois Michonneau}\\
-\noindent {\small \it $^1$Florida Museum of Natural History \& Whitney Laboratory
-  for Marine Sciences, University of Florida, Gainesville,
-  FL 32611-7800, USA}\\
+\noindent{\normalsize \sc Fran\c{c}ois Michonneau$^1$, Joseph Brown$^2$, David Winter$^3$}\\
+\noindent {\small \it $^1$Florida Museum of Natural History \& Whitney
+  Laboratory for Marine Sciences, University of Florida, Gainesville,
+  FL, USA}\\
+\noindent{\small \it $^2$University of Michigan, Ann Arbor, MI, USA} \\
+\noindent{\small \it $^3$Arizona State University, Tempe, AZ, USA} \\
 \end{center}

### 2015-10-09 manuscript.Rnw
improve figure

@@
 
-<<plot_taxa, cache=TRUE, fig.cap="The phylogenetic tree returned by OTL for the list of model species used as an example.">>=
+<<plot_taxa, fig.height=4, cache=TRUE, fig.cap="The phylogenetic tree returned by OTL for the list of model species used as an example.">>=
 tree <- tol_induced_subtree(ott_ids = taxa[["ott_id"]])
-plot(tree)
+plot(tree, cex=.8, label.offset=.1, no.margin=TRUE)
 @
@@
 print(cat_tree, printlen = 4)
+@
+
+<<plot_cats, fig.height=4.2, echo=FALSE, fig.cap='Phylogeny of the Felidae published in \\citealt{Johnson2006} and retrieved from OTL using \\texttt{rotl}'>>=
+plot(cat_tree, label.offset=.0001, cex=.75, no.margin=TRUE)
 @
@@
     coord_map(projection = "mercator", orientation = c(90, 0, 0)) +
-    xlab("Longitude") + ylab("Latitude")
+    xlab("Longitude") + ylab("Latitude") +
+    theme(legend.position="top") + ylim(c(0,72))
 

### 2015-10-09 manuscript.Rnw
write availability section

@@
 
-\texttt{rotl} is available both from the CRAN repository
-(https://cran.r-project.org/package=rotl) and from GitHub
-(https://github.com/ropensci/rotl).
+\texttt{rotl} is free, open source, and released under a Simplified
+BSD license. Stable versions are available from the CRAN repository
+\url{https://cran.r-project.org/package=rotl}, and development
+versions are available from GitHub
+\url{https://github.com/ropensci/rotl}.
+
+Python \url{https://github.com/OpenTreeOfLife/pyopentree} and Ruby
+\url{https://github.com/SpeciesFileGroup/bark} libraries to interact
+with the OTL APIs are also available.
 

### 2015-10-09 manuscript.Rnw
various edits

@@
 2.3 million tips. Beyond obvious applications across the life sciences to
-explore research questions in evolution, biodiversity, and conservation, the
-resources OTL provides could be used for education and outreach (e.g.,
-illustrating course material, develop outreach activities to explore
-relationships among species).
+explore questions in evolution, biodiversity, and conservation, the resources
+OTL provides could be used for education and outreach (e.g., illustrating course
+material, develop outreach activities to explore relationships among species).
 
@@
 download parts of the synthetic Open Tree
-(\url{http://tree.opentreeoflife.org}). Additionally, the project provides an
-Application Programming Interface (API) to the different services to work
+(\url{http://tree.opentreeoflife.org}). Additionally, the project provides
+Application Programming Interfaces (APIs) to the different services to work
 programmatically with the underlying data. Here, we present \texttt{rotl} an R
-package to interact with OTL's API using R syntax to search for, and import
+package to interact with OTL's APIs using R syntax to search for, and import
 phylogenetic trees and taxonomic information into R. Having access to this data
@@
 are available to visualize, analyze and manipulate phylogenetic and comparative
-data (ape, phytools, geiger, ggtree, phylobase, RNeXML, etc... cite task view
-\url{https://cran.r-project.org/web/views/Phylogenetics.html}).
+data (e.g., ape, phytools, geiger, ggtree, phylobase, RNeXML, see
+\url{https://cran.r-project.org/web/views/Phylogenetics.html} for a
+comprehensive list). Furthermore, \texttt{rotl} joins a growing list of R
+packages that aim at accessing data from the web.
 
-\section{Services provided by OTL}
+\section{API services provided by OTL}
 
 The OTL project provides four resources that serve data to users through the
-API:
+APIs:
 
@@
   names to the Open Tree Taxonomy identifiers;
-\item The \emph{synthetic tree} itself (the Open Tree).
+\item The \emph{synthetic tree} itself, the Open Tree.
 \end{enumerate}
@@
 Additionally, the project provides access to other data, the Graph of Life, that
-stores information about the underlying tree alignment graphs \cite{Smith2013}
+stores information about the underlying tree alignment graphs \citep{Smith2013}
 used to synthesize the various sources of information from which the Open Tree
-is assembled from. Data from this API endpoint is not intended for normal users
-and its use will not be demonstrated in this paper but \texttt{rotl} has
-functions to retrieve data it provides.
+is assembled from. Data from this API is not intended for normal users, and its
+use will not be demonstrated in this paper but \texttt{rotl} has functions to
+retrieve data it provides.
 
@@
 it, therefore parametric comparative methods cannot be used directly on the
-trees returned by the API. However, resources and methods are being developed to
-add branch lengths information to these topological trees
+trees returned by OTL. However, resources and methods are being developed to add
+branch lengths information to these topological trees
 \citep[e.g.,][]{Ksepka2015} or use topological trees to identify
@@
 \url{https://cran.r-project.org/package=rncl}. Using NCL provides robust and
-efficient parsing of large trees that may contain singletons nodes labeled with
+efficient parsing of large trees that may contain singleton nodes labeled with
 taxonomic information.
@@
 replaces it in the results. An example of this approach is provided in the
-vignette ``How to use rotl?'' that accompanies the package.
+vignette ``How to use \texttt{rotl}?'' that accompanies the package.
 
@@
 \texttt{list\_trees} returns a list containing the tree identifiers for each
-study easily. Alternatively, the function \texttt{get\_study} returns all the
-trees (by default as \texttt{ape::phylo} objects) associated with a particular
+study. Alternatively, the function \texttt{get\_study} returns all the trees (by
+default as \texttt{ape::phylo} objects) associated with a particular
 study. Additional details about the study (i.e., its metadata) can be obtained
@@
 data from OTL can be combined with other sources easily. For instance in
-Figure~\ref{cat_map}, we show how we can obtain a map of the occurrences for
+Figure~\ref{fig:cat_map}, we show how we can obtain a map of the occurrences for
 some of the cat species (genus \textit{Lynx}) included in the phylogeny above
@@
 
-<<cat_map, echo=FALSE, results='hide', message=FALSE, cache=TRUE, fig.cap='GBIF records for the species in \\textit{Lynx} included in the phylogeny associated with the study by \\citealt{Johnson2006}'>>=
+<<cat_map, fig.height=4.2, echo=FALSE, results='hide', message=FALSE, cache=TRUE, fig.cap='GBIF records for the species in \\textit{Lynx} included in the phylogeny associated with the study by \\citealt{Johnson2006}'>>=
 
@@
 lynx_loc[["decimalLongitude"]] <- as.numeric(lynx_loc[["decimalLongitude"]])
+lynx_loc[lynx_loc[["decimalLatitude"]] == 0 &
+         lynx_loc[["decimalLongitude"]] == 0,
+         c("decimalLatitude","decimalLongitude")] <- c(NA, NA)
 lynx_loc <- lynx_loc[complete.cases(lynx_loc), ]

### 2015-10-15 manuscript.Rnw
added citations for R packages listed in intro

@@
 (\url{http://tree.opentreeoflife.org}). Additionally, the project provides
-Application Programming Interfaces (APIs) to the different services to work
-programmatically with the underlying data. Here, we present \texttt{rotl} an R
-package to interact with OTL's APIs using R syntax to search for, and import
-phylogenetic trees and taxonomic information into R. Having access to this data
-directly in R allows researchers to take advantage of the multiple packages that
-are available to visualize, analyze and manipulate phylogenetic and comparative
-data (e.g., ape, phytools, geiger, ggtree, phylobase, RNeXML, see
+Application Programming Interfaces (APIs) to access programmatically the
+different services and underlying data. Here, we present \texttt{rotl} a package
+to interact with OTL's APIs to search for and import phylogenetic trees and
+taxonomic information directly in R. Having access to this data in R allows
+researchers to take advantage of the multiple packages that are available to
+visualize, analyze and manipulate phylogenetic and comparative data (e.g., ape
+\citealt{Paradis2004}, phytools \citealt{Revell2012}, geiger
+\citealt{Pennell2014-geiger}, ggtree \citealt{Yu-ggtree}, phylobase
+\citealt{phylobase-0.8.0}, RNeXML \citealt{Boettiger2015}, see
 \url{https://cran.r-project.org/web/views/Phylogenetics.html} for a
-comprehensive list). Furthermore, \texttt{rotl} joins a growing list of R
-packages that aim at accessing data from the web.
+comprehensive list). Furthermore, \texttt{rotl} joins a growing list of packages
+that aim at querying and accessing data from the web directly in R for analysis.
 

### 2015-10-15 manuscript.Rnw
small edits to the text

@@
 efficient parsing of large trees that may contain singleton nodes labeled with
-taxonomic information.
+taxonomic information (i.e., monotypic taxon).
 
@@
 The function \texttt{tnrs\_match\_names} returns a data frame that lists the
-Open Tree identifiers as well as other information to help users to ensure that
-the taxa matched are the correct ones. Here, there is no ambiguity in the taxa
+Open Tree identifiers as well as other information to help users ensure that the
+taxa matched are the correct ones. Here, there is no ambiguity in the taxa
 matched, however, as the Open Tree Taxonomy includes taxa from bacteria, plants,
@@
 hemihomonyms. The argument \texttt{context\_name} can be used to limit potential
-matches to a taxonomic group (see the function \texttt{tnrs\_contexts} for a list
-of possible options).  When this strategy cannot be used in case the tree
-encompasses multiple domains (as in the present example), the function
-\texttt{inspect} lists alternative matches for a taxon name, and \texttt{update}
-replaces it in the results. An example of this approach is provided in the
-vignette ``How to use \texttt{rotl}?'' that accompanies the package.
+matches to a taxonomic group such as ``Animals'' (see the function
+\texttt{tnrs\_contexts} for a list of possible options).  When this strategy
+cannot be used in case the tree encompasses multiple domains (as in the present
+example), the function \texttt{inspect} lists alternative matches for a taxon
+name, and \texttt{update} replaces it in the results. An example of this
+approach is provided in the vignette ``How to use \texttt{rotl}?'' that
+accompanies the package.
 
@@
 criteria such as the title of the publication can be used. Here we demonstrate
-how to look for and retrieve a tree for study focusing on the family Felidae
+how to look for and retrieve a tree for studies focusing on the family Felidae
 (Figure~\ref{fig:plot_cats}).
@@
 methods in novel ways. Additionally, as more data is made available online and
-accessible using APIs, many R packages have been developed to take advantage of
+accessible using APIs, several R packages have been developed to take advantage of
 these datasets, especially spearheaded by the organization rOpenSci

### 2015-10-15 manuscript.Rnw
add comments in script to build Lynx map

@@
 <<cat_map, fig.height=4.2, echo=FALSE, results='hide', message=FALSE, cache=TRUE, fig.cap='GBIF records for the species in \\textit{Lynx} included in the phylogeny associated with the study by \\citealt{Johnson2006}'>>=
-
+## load required packages
 suppressPackageStartupMessages(library(rgbif))
@@
 
+## Find the species of Lynx in the phylogeny
 cat_species <- cat_tree$tip.label
@@
 
+## Match the Lynx species to the GBIF identifiers
 gbif_keys <- sapply(lynx_species,
@@
 
+## Search for the GBIF records for these species
 lynx_loc <- occ_search(taxonKey = gbif_keys, limit = 500,
@@
 
+## Make a data frame of the results, and clean up the data with missing locality data
 lynx_loc <- do.call("rbind", lynx_loc)
@@
 
+## Draw the map
 world <- map_data("world")

### 2015-10-16 manuscript.Rnw
add abstract

@@
 
-Abstract goes here.
+While phylogenies have been getting easier to build, it has been difficult to
+re-use, combine and synthesize the information they provide because published
+trees are often only available as image files, and taxonomic information is not
+standardized across studies. The Open Tree of Life (OTL) project addresses these
+issues by providing a digital tree that encompasses all organisms built by
+combining taxonomic information and published phylogenies. The project also
+provides tools and services to query and download parts of this synthetic tree,
+as well as the data used to build it. Here, we present \texttt{rotl}, an R
+package to search and download data from the Open Tree of Life data directly in
+R. \texttt{rotl} uses common data structures allowing researchers to take
+advantage of the rich set of tools and methods that are available in R to
+manipulate, analyze and visualize phylogenies.
 

### 2015-10-16 manuscript.Rnw
better affiliations

@@
 
-\noindent{\normalsize \sc Fran\c{c}ois Michonneau$^1$, Joseph Brown$^2$, David Winter$^3$}\\
-\noindent {\small \it $^1$Florida Museum of Natural History \& Whitney
-  Laboratory for Marine Sciences, University of Florida, Gainesville,
-  FL, USA}\\
-\noindent{\small \it $^2$University of Michigan, Ann Arbor, MI, USA} \\
-\noindent{\small \it $^3$Arizona State University, Tempe, AZ, USA} \\
+\noindent{\normalsize \sc Fran\c{c}ois Michonneau$^{1,2}$, Joseph Brown$^3$, David Winter$^4$}\\
+
+\noindent {\small \it $^1$ Whitney Laboratory for Marine Sciences, University of
+  Florida, St. Augustine, FL, USA}\\
+\noindent {\small \it $^2$Florida Museum of Natural History, University of
+  Florida, Gainesville, FL, USA} \\
+\noindent{\small \it $^3$University of Michigan, Ann Arbor, MI, USA} \\
+\noindent{\small \it $^4$Arizona State University, Tempe, AZ, USA} \\
 \end{center}

### 2015-10-16 manuscript.Rnw
add conclusion

@@
 
-\section{Conclusion}
+\section{Concluding remarks}
 \label{sec:conclusion}
 
-Resource as useful as the data that is in it. To make data available, it needs
-to follow conventions so it can be incorporated into the OTL, i.e., submit tree
-files (not images) to public repositories.
+The recognition of the importance of phylogenies to account for the statistical
+non-independence of species in comparative methods, and the recent development
+of methods to explore trait evolution or changes in diversification rates have
+driven the need for accurate phylogenies. However, there is often a discrepancy
+between taxa targeted by comparative methods, and taxa for which phylogenies are
+available. We believe that by providing an easy-to-use interface to obtain
+phylogenies for an arbitrary set of taxa directly in R, \texttt{rotl} will be
+useful in a wide variety of contexts.
 
-OTL provides an easy-to-use interface to submit studies.
+The accuracy and usefulness of the data provided by OTL relies on the community
+to make generated phylogenies (and their metadata) digitally available as tree
+files (i.e., Newick, NEXUS or NeXML). We strongly encourage researchers to
+submit their published phylogenies to OTL using the curator interface
+\url{https://tree.opentreeoflife.org/curator}. By facilitating the discovery and
+re-use of published trees and of the synthetic tree, we hope \texttt{rotl} will
+contribute to the wider adoption of best practices to make phylogenetic
+information available and re-usable.
 

### 2015-10-16 manuscript.Rnw
add acknowledgments

@@
 
-* OpenTree of Life hackathon
-* Scott Chamberlain for code review.
+We would like to thank the organizers of the OpenTree of Life APIs hackathon
+that was held at the University of Michigan, Ann Arbor, September 15-19, 2014,
+where the development of \texttt{rotl} was started. We would also like to thank
+Scott Chamberlain (rOpenSci) for providing a thorough code review. FM was
+supported by iDigBio, and therefore this material is based upon work supported
+by the National Science Foundation’s Advancing Digitization of Biodiversity
+Collections Program (Cooperative Agreement EF-1115210)
 

### 2015-10-16 manuscript.Rnw
various edits

@@
 morphological and ecological traits, the spatio-temporal variation in speciation
-rates, or both \citep{Pennell2013}. The development of these methods has
-contributed to the need for a greater availability of high-quality phylogenetic
-information across life. However, phylogenetic information is scattered, often
-only available as image files within publications, and the variety of file
-formats used to store phylogenetic data makes it difficult for researchers to
-access, synthesize, and integrate this information into their own research
-(\citealt{Stoltzfus2012}, but see \citealt{Cranston2014} for suggestions of best
-practices).
+rates, or both \citep{OMeara2012,Pennell2013}. The development of these methods
+has contributed to the need for a greater availability of high-quality
+phylogenetic information across life. However, phylogenetic information is
+scattered, often only available as image files within publications, and the lack
+of standardization to store and represent phylogenetic data makes it difficult
+for researchers to access, synthesize, and integrate this information into their
+own research (\citealt{Stoltzfus2012}, but see \citealt{Cranston2014} for
+suggestions of best practices).
 
@@
 trees returned by OTL. However, resources and methods are being developed to add
-branch lengths information to these topological trees
-\citep[e.g.,][]{Ksepka2015} or use topological trees to identify
-phylogenetically equivalent species to increase overlap between chronograms and
-species trait data \citep{Pennell2015}. Without branch lengths, these trees
-remain useful to illustrate relationships among species, or to map a trait on a
-phylogeny for instance.
+branch lengths to these topological trees \citep[e.g.,][]{Ksepka2015} or use
+topological trees to identify phylogenetically equivalent species to increase
+overlap between chronograms and species trait data \citep{Pennell2015}. Without
+branch lengths, these trees are nonetheless useful to illustrate relationships
+among species, or to map a trait on a phylogeny for instance.
 
@@
 need to be matched against the Open Taxonomy using the TNRS. This step retrieves
-the taxa identifiers that will be used to extract the relationships for the set
-of requested taxa.
+the identifiers that will be used to extract the relationships for the set of
+requested taxa.
 
@@
 contribute to the synthetic tree. These trees provide a useful resource to
-reproduce or expand on an already published analysis.
+reproduce or expand on an already published analysis, or to explore how the
+elucidation of relationships within a clade has changed through time.
 
@@
 
-<<get_cats, fig.cap="Tree from \\cite{} obtained from OTL">>=
+<<get_cats, purl=FALSE, fig.cap="Tree from \\citealt{Johnson2006} obtained from OTL">>=
 cat_studies <- studies_find_studies(property = "ot:focalCladeOTTTaxonName",
@@
 In recent years, R has become part of the toolbox of many researchers in
-evolutionary biology. As many methods for comparative analyses are implemented,
-and because it is a relatively easy to use programming language, R greatly
-facilitates the analysis of large datasets, and allows researchers to combine
-methods in novel ways. Additionally, as more data is made available online and
-accessible using APIs, several R packages have been developed to take advantage of
-these datasets, especially spearheaded by the organization rOpenSci
-\url{https://ropensci.org}. \texttt{rotl} contributes to these initiatives and
-data from OTL can be combined with other sources easily. For instance in
-Figure~\ref{fig:cat_map}, we show how we can obtain a map of the occurrences for
-some of the cat species (genus \textit{Lynx}) included in the phylogeny above
-using records for these species found in GBIF (code is available in
-supplementary material).
+evolutionary biology. R greatly facilitates the analysis of large datasets, and
+allows researchers to combine methods in novel ways because many methods for
+comparative analyses are implemented, and because it is a relatively easy to use
+programming language. Additionally, as more data is made available online and
+accessible using APIs, several packages have been developed to interact and
+download these datasets directly in R enabling direct and reproducible analyses.
+Notably, the organization rOpenSci \url{https://ropensci.org} has fostered a
+community of researchers who develop tools and methods to facilitate the use of
+open data and broaden the adoption of open science practices in
+general. \texttt{rotl} contributes to this initiative, and data from OTL can be
+combined with other sources easily. For instance in Figure~\ref{fig:cat_map}, we
+show how we can obtain a map of the occurrences for some of the cat species
+(genus \textit{Lynx}) included in the phylogeny above using records for these
+species found in GBIF (code available in Appendix). Another example is available
+in the package vignette ``Data mashups''.
 
-<<cat_map, fig.height=4.2, echo=FALSE, results='hide', message=FALSE, cache=TRUE, fig.cap='GBIF records for the species in \\textit{Lynx} included in the phylogeny associated with the study by \\citealt{Johnson2006}'>>=
+<<cat_map, purl=TRUE, echo=FALSE, size='small', fig.height=4.2, results='hide', message=FALSE, cache=TRUE, fig.cap='GBIF records for the species in \\textit{Lynx} included in the phylogeny associated with the study by \\citealt{Johnson2006}'>>=
 ## load required packages
@@
 versions are available from GitHub
-\url{https://github.com/ropensci/rotl}.
+\url{https://github.com/ropensci/rotl}. The source for this manuscript is
+available at \url{https://github.com/fmichonneau/rotl-ms}.
 

### 2015-10-16 manuscript.Rnw
create standalone appendix from main manuscript

@@
 
-<<init, echo=FALSE>>=
+<<init, echo=FALSE, purl=FALSE>>=
 library(rotl)
@@
 
-<<get_taxa, cache=TRUE>>=
+<<get_taxa, cache=TRUE, purl=FALSE>>=
 taxa <- tnrs_match_names(names = c("Escherichia colli",
@@
 
-<<plot_taxa, fig.height=4, cache=TRUE, fig.cap="The phylogenetic tree returned by OTL for the list of model species used as an example.">>=
+<<plot_taxa, fig.height=4, cache=TRUE, purl=FALSE, fig.cap="The phylogenetic tree returned by OTL for the list of model species used as an example.">>=
 tree <- tol_induced_subtree(ott_ids = taxa[["ott_id"]])
@@
 
-<<get_cat_tree, results='hide'>>=
+<<get_cat_tree, purl=FALSE, results='hide'>>=
 cat_tree <- get_study_tree(study_id = cat_studies[["study_ids"]][1],
@@
 
-<<print_cat_tree, echo=FALSE>>=
+<<print_cat_tree, purl=FALSE, echo=FALSE>>=
 print(cat_tree, printlen = 4)
@@
 
-<<plot_cats, fig.height=4.2, echo=FALSE, fig.cap='Phylogeny of the Felidae published in \\citealt{Johnson2006} and retrieved from OTL using \\texttt{rotl}'>>=
+<<plot_cats, purl=FALSE, fig.height=4.2, echo=FALSE, fig.cap='Phylogeny of the Felidae published in \\citealt{Johnson2006} and retrieved from OTL using \\texttt{rotl}'>>=
 plot(cat_tree, label.offset=.0001, cex=.75, no.margin=TRUE)
@@
 ## load required packages
-suppressPackageStartupMessages(library(rgbif))
-suppressPackageStartupMessages(library(ggplot2))
-suppressPackageStartupMessages(library(maps))
+library(rgbif)
+library(ggplot2)
+library(maps)
 
@@
 
-## Make a data frame of the results, and clean up the data with missing locality data
+## Make a data frame of the results
 lynx_loc <- do.call("rbind", lynx_loc)
+names(lynx_loc)[1] <- "Species"
+
+## Clean up the data with missing locality data
 lynx_loc[["decimalLatitude"]] <- as.numeric(lynx_loc[["decimalLatitude"]])
@@
 lynx_loc <- lynx_loc[complete.cases(lynx_loc), ]
-
-names(lynx_loc)[1] <- "Species"
 

### 2015-10-16 create_appendix.R

@@ -1,0 +1,18 @@
+knitr::purl("manuscript.Rnw", documentation = 0L)
+
+header <- c("---",
+            "title: Appendix",
+            "output: pdf_document",
+            "---",
+            "\n",
+            "Source code used to generate Figure 3.",
+            "```{r, echo=TRUE, eval=FALSE}")
+
+footer <- "```"
+
+code_chunk <- readLines("manuscript.R")
+
+cat(header, code_chunk, footer, sep = "\n",
+    file = "manuscript-appendix.Rmd")
+
+rmarkdown::render("manuscript-appendix.Rmd")

### 2015-10-21 manuscript.Rnw
DJW edit
Mostly to intro and pretty minimal
 * Slightly more detail on growth of phylo, lack of synth. at present
 * Increased emph. on R/rotl as reproducable research
 * Mention published trees (not just synth) in inro

Other minor changes
 * Added comment about accepting features/bugs in availability
 * Short blurb about tests/docs in technical details

@@
 
-%%%%% packages provided by sys bio template
-\usepackage{fixltx2e}
-\usepackage{textcomp}
-\usepackage{fullpage}
-\usepackage{amsfonts}
-\usepackage{verbatim}
-%% \usepackage[english]{babel} %% I use polyglossia instead
-%% not needed: \usepackage{pifont}
-\usepackage{color}
-\usepackage{setspace}
-\usepackage{lscape}
-\usepackage{indentfirst}
-\usepackage[normalem]{ulem}
-\usepackage{booktabs}
-%\usepackage{nag}
-\usepackage{natbib}
-%\usepackage{bibtex}
-\usepackage{float}
-\usepackage{latexsym}
-%\usepackage{hyperref}
-\usepackage{url}
-%\usepackage{html}
-\usepackage{hyperref}
-\usepackage{epsfig}
-\usepackage{graphicx}
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{bm}
-\usepackage{array}
-%\usepackage{mhchem}
-\usepackage{ifthen}
-%% I am already using it: \usepackage{caption}
-\usepackage{hyperref}
-%\usepackage{xcolor}
-\usepackage{amsthm}
-\usepackage{amstext}
-%%%%
+%%%%% packages provided by sys bio template \usepackage{fixltx2e}
+\usepackage{textcomp} \usepackage{fullpage} \usepackage{amsfonts}
+\usepackage{verbatim} %% \usepackage[english]{babel} %% I use polyglossia
+instead %% not needed: \usepackage{pifont} \usepackage{color}
+\usepackage{setspace} \usepackage{lscape} \usepackage{indentfirst}
+\usepackage[normalem]{ulem} \usepackage{booktabs} %\usepackage{nag}
+\usepackage{natbib} %\usepackage{bibtex} \usepackage{float}
+\usepackage{latexsym} %\usepackage{hyperref} \usepackage{url} %\usepackage{html}
+\usepackage{hyperref} \usepackage{epsfig} \usepackage{graphicx}
+\usepackage{amssymb} \usepackage{amsmath} \usepackage{bm} \usepackage{array}
+%\usepackage{mhchem} \usepackage{ifthen} %% I am already using it:
+\usepackage{caption} \usepackage{hyperref} %\usepackage{xcolor}
+\usepackage{amsthm} \usepackage{amstext} %%%%
 
 % line numbers
-\usepackage{lineno}
-\modulolinenumbers[5]
-\linenumbers
+\usepackage{lineno} \modulolinenumbers[5] \linenumbers
 
-\usepackage{natbib}
-\usepackage[T1]{fontenc}
-\usepackage{graphicx}
-\usepackage[margin=1in]{geometry}
-\usepackage[font=sf]{caption}
+\usepackage{natbib} \usepackage[T1]{fontenc} \usepackage{graphicx}
+\usepackage[margin=1in]{geometry} \usepackage[font=sf]{caption}
 
 % Latex special characters are rendered correctly with XeTeX
-\usepackage{xltxtra}
-\usepackage{xunicode}
+\usepackage{xltxtra} \usepackage{xunicode}
 \defaultfontfeatures{Mapping=tex-text}
@@
 % Words are cut where needed
-\usepackage{polyglossia}
-\setdefaultlanguage[variant=american]{english}
+\usepackage{polyglossia} \setdefaultlanguage[variant=american]{english}
 
 % Use fancy fonts
-\usepackage{fontspec}
-\setmainfont[Mapping=tex-text]{Crimson Text}
+\usepackage{fontspec} \setmainfont[Mapping=tex-text]{Crimson Text}
 \setsansfont{SourceSansPro-Regular}
 
-%%% syst bio options
-\linespread{1.66}
-% All text should be double-spaced
-% with occasional exceptions for tables.
-\raggedright
-\setlength{\parindent}{0.5in}
+%%% syst bio options \linespread{1.66}
+% All text should be double-spaced with occasional exceptions for tables.
+\raggedright \setlength{\parindent}{0.5in}
 
@@
 
-% Our sections are not numbered and our papers do not have
-% Tables of Contents. We don't
-% present a list of figures or list of tables, either.
+% Our sections are not numbered and our papers do not have Tables of Contents.
+% We don't present a list of figures or list of tables, either.
 
-% Any common font is fine.
-% (A common sans-serif font should be used on figures, but figures should be
-% separate from the LaTeX document.)
+% Any common font is fine.  (A common sans-serif font should be used on figures,
+% but figures should be separate from the LaTeX document.)
 
@@
 
-\renewcommand{\section}[1]{%
-\bigskip
-\begin{center}
-\begin{Large}
-\normalfont\scshape #1
-\medskip
-\end{Large}
-\end{center}}
+\renewcommand{\section}[1]{% \bigskip \begin{center} \begin{Large}
+        \normalfont\scshape #1 \medskip \end{Large} \end{center}}
 
-\renewcommand{\subsection}[1]{%
-\bigskip
-\begin{center}
-\begin{large}
-\normalfont\itshape #1
-\end{large}
-\end{center}}
+\renewcommand{\subsection}[1]{% \bigskip \begin{center} \begin{large}
+        \normalfont\itshape #1 \end{large} \end{center}}
 
-\renewcommand{\subsubsection}[1]{%
-\vspace{2ex}
-\noindent
-\textit{#1.}---}
+\renewcommand{\subsubsection}[1]{% \vspace{2ex} \noindent \textit{#1.}---}
 
@@
 
-\bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib
-%%% end syst bio options
+\bibpunct{(}{)}{;}{a}{}{,}  % this is a citation format command for natbib %%%
+end syst bio options
 
-\newcolumntype{b}{X}
-\newcolumntype{s}{>{\hsize=.2\hsize}X}
+\newcolumntype{b}{X} \newcolumntype{s}{>{\hsize=.2\hsize}X}
 
-% ----------------------------------------------------------------------------- %
+% -----------------------------------------------------------------------------
+% %
 
@@
 
-\begin{flushright}
-Version dated: \today
-\end{flushright}
+\begin{flushright} Version dated: \today \end{flushright}
 
-\bigskip
-\noindent AN R PACKAGE TO INTERACT WITH THE OPEN TREE OF LIFE
+\bigskip \noindent AN R PACKAGE TO INTERACT WITH THE OPEN TREE OF LIFE
 
-\bigskip
-\medskip
+\bigskip \medskip
 
-\begin{center}
-  \noindent{\Large \bf \texttt{rotl} an R package to interact with the Open Tree
-    of Life Data}
+\begin{center} \noindent{\Large \bf \texttt{rotl} an R package to interact with
+the Open Tree of Life Data}
 
@@
 
-\noindent{\normalsize \sc Fran\c{c}ois Michonneau$^{1,2}$, Joseph Brown$^3$, David Winter$^4$}\\
+\noindent{\normalsize \sc Fran\c{c}ois Michonneau$^{1,2}$, Joseph Brown$^3$,
+David Winter$^4$}\\
 
 \noindent {\small \it $^1$ Whitney Laboratory for Marine Sciences, University of
-  Florida, St. Augustine, FL, USA}\\
-\noindent {\small \it $^2$Florida Museum of Natural History, University of
-  Florida, Gainesville, FL, USA} \\
+Florida, St. Augustine, FL, USA}\\ \noindent {\small \it $^2$Florida Museum of
+Natural History, University of Florida, Gainesville, FL, USA} \\
 \noindent{\small \it $^3$University of Michigan, Ann Arbor, MI, USA} \\
 \noindent{\small \it $^4$Arizona State University, Tempe, AZ, USA} \\
-\end{center}
-\medskip
+\end{center} \medskip
 
@@
 
-Generating phylogenies has been greatly facilitated by advances in sequencing
-and computing technologies. In parallel, new phylogenetic comparative methods
-have been developed and make use of these phylogenies to explore fundamental
-questions about the origin of biodiversity including the evolution of
-morphological and ecological traits, the spatio-temporal variation in speciation
-rates, or both \citep{OMeara2012,Pennell2013}. The development of these methods
-has contributed to the need for a greater availability of high-quality
-phylogenetic information across life. However, phylogenetic information is
-scattered, often only available as image files within publications, and the lack
-of standardization to store and represent phylogenetic data makes it difficult
-for researchers to access, synthesize, and integrate this information into their
-own research (\citealt{Stoltzfus2012}, but see \citealt{Cranston2014} for
-suggestions of best practices).
+Advances in sequencing and computing technologies have lead to a revolution in
+systematic biology. The ability to routinely generate molecular datasets from
+any extant organism has allowed researchers to resolve long-running taxonomic
+disputes and estimate phylogenies for previously understudied groups. In
+parallel, the ease which which phylogenies can be estimated has lead to the
+development of new phylogenetic comparative methods. These methods allow
+researchers to explore fundamental questions about the origin of biodiversity
+including the evolution of morphological and ecological traits, the
+spatio-temporal variation in speciation rates, or both
+\citep{OMeara2012,Pennell2013}.
+
+Ideally, the ever increasing number of published phylogenies would contribute to
+a synthesis of phylogenetic knowledge -- leading to a better understanding of
+the history of life and providing high-quality phylogenetic information for use
+in comparative methods. However, in practice synthesizing phylogenetic data
+remains difficult. Phylogenies are scattered, often only available as image
+files within publications, and the lack of standardization to store and
+represent phylogenetic data makes it difficult for researchers to access,
+synthesize, and integrate this information into their own research
+(\citealt{Stoltzfus2012}, but see \citealt{Cranston2014} for suggestions of best
+practices).
 
@@
 The Open Tree of Life project provides a web interface to explore, query and
-download parts of the synthetic Open Tree
-(\url{http://tree.opentreeoflife.org}). Additionally, the project provides
-Application Programming Interfaces (APIs) to access programmatically the
-different services and underlying data. Here, we present \texttt{rotl} a package
-to interact with OTL's APIs to search for and import phylogenetic trees and
-taxonomic information directly in R. Having access to this data in R allows
+download parts of the synthetic Open Tree (\url{http://tree.opentreeoflife.org})
+and the published trees that contribute to it. Additionally, the project
+provides Application Programming Interfaces (APIs)  to access programmatically
+the different services and underlying data.  Here, we present \texttt{rotl} a
+package to interact with OTL's APIs to search for and import phylogenetic trees
+and taxonomic information directly in R. Having access to this data in R allows
 researchers to take advantage of the multiple packages that are available to
 visualize, analyze and manipulate phylogenetic and comparative data (e.g., ape
-\citealt{Paradis2004}, phytools \citealt{Revell2012}, geiger
+\citealt{Paradis2004}, phytools \citeateu{Revell2012}, geiger
 \citealt{Pennell2014-geiger}, ggtree \citealt{Yu-ggtree}, phylobase
@@
 \url{https://cran.r-project.org/web/views/Phylogenetics.html} for a
-comprehensive list). Furthermore, \texttt{rotl} joins a growing list of packages
-that aim at querying and accessing data from the web directly in R for analysis.
+comprehensive list). There is also a growing list of packages that allow users
+to query and acces data from the web directly in R \hl{rFISHBASE
+10.1111/j.1095-8649.2012.03464.x, paleoDB 10.1111/ecog.01154, rAvis
+10.1371/journal.pone.0091650 \dots}. By providing direct access to high quality
+phylogenetic data in R, \texttt{rotl} fills a key gap typical comparative
+analysis workflows. Moreover, by allowing the gathering, processing and analysis
+of data to be performed programmatically, \texttt{rotl} facilitates reproducible
+research practices.
 
@@
 taxonomic information (i.e., monotypic taxon).
+
+The package is well-documented, and includes two package
+vignettes (documents that demonstrate the use of the package and contain executable 
+R code) as well as per-function documentation. There is also an extensive
+test-suite that covers both the internal functions that \textt{rotl} uses to
+connect to OTL and public functions users apply to acess and process data.
 
@@
 versions are available from GitHub
-\url{https://github.com/ropensci/rotl}. The source for this manuscript is
-available at \url{https://github.com/fmichonneau/rotl-ms}.
+\url{https://github.com/ropensci/rotl}. The package is under active development,
+and authors welcome bug reports of feature requests via the github repository.
+The source for this manuscript is available at \url{https://github.com/fmichonneau/rotl-ms}.
 

### 2015-10-26 manuscript.Rnw
DJW edits to intro/ecosystem sections
 * reworded final intro sentence -- emphasis on R as platform for repro. research
 * adds citations to data packages/ knitr
 * added ref. to rTreeBase in 'ecoystem' section
 * descirption of meta-analysis vignette
 * reword intro to cat_map to match other edits to paragraph

@@
 %%%%
+
+%% Highlighting 
+\usepackage{soul}
 
@@
 
-The Open Tree of Life project provides a web interface to explore, query and
-download parts of the synthetic Open Tree (\url{http://tree.opentreeoflife.org})
-and the published trees that contribute to it. Additionally, the project
-provides Application Programming Interfaces (APIs)  to access programmatically
-the different services and underlying data.  Here, we present \texttt{rotl} a
-package to interact with OTL's APIs to search for and import phylogenetic trees
-and taxonomic information directly in R. Having access to this data in R allows
-researchers to take advantage of the multiple packages that are available to
-visualize, analyze and manipulate phylogenetic and comparative data (e.g., ape
-\citealt{Paradis2004}, phytools \citeateu{Revell2012}, geiger
-\citealt{Pennell2014-geiger}, ggtree \citealt{Yu-ggtree}, phylobase
-\citealt{phylobase-0.8.0}, RNeXML \citealt{Boettiger2015}, see
+The R programming language is a popular tool for  phylogenetics and comparative
+analysis. The R packages \texttt{ape} \cite{Paradis2004}, \texttt{phylobase}
+\cite{phylobase-0.8.0 }and \texttt{RNeXML} \cite{Boettiger2015} provide
+functions to import and manipulate trees within R and save the results in
+standard data formats. Additional packages including  \texttt{phytools}
+\cite{Revell2012}, \texttt{geiger} \cite{Pennell2014-geiger} and \texttt{ggtree} 
+\cite{Yu-ggtree} to allow users to analyse and visualise data in a 
+phylogenetic context (see 
 \url{https://cran.r-project.org/web/views/Phylogenetics.html} for a
-comprehensive list). There is also a growing list of packages that allow users
-to query and acces data from the web directly in R \hl{rFISHBASE
-10.1111/j.1095-8649.2012.03464.x, paleoDB 10.1111/ecog.01154, rAvis
-10.1371/journal.pone.0091650 \dots}. By providing direct access to high quality
+comprehensive list of phylogenetics packages in R). In addition to packages for
+phylogentic analyses, a growing number of R packages allow users to query and
+acess data from the web \hl{(rFISHBASE 10.1111/j.1095-8649.2012.03464.x, 
+paleoDB 10.1111/ecog.01154,  rAvis 10.1371/journal.pone.0091650)}, meaning data 
+associated with taxa in a given phylogeny can be obtained directly in R. 
+These packages, combined the language's support for literatre programming 
+\hl{(Knuth\_1984, Xie\_2014)} make R an ideal platform for reproducable research
+in phylogentics and comparative biology, as they allow a complete record of the 
+steps taken in gathering, processing and analysing a given data set to be
+produced.
+
+Here we present \texttt{rotl}, and R package that allows users to download
+phylogentic and taxonomic data from Open Tree of Life in R. \texttt{rotl} takes
+adavantage of Application Programming Interfaces (APIs) provided by OTL to acess
+sub-trees from the synthetic Open Tree and the published trees that contribute
+to the synthesis. By providing direct access to high quality
 phylogenetic data in R, \texttt{rotl} fills a key gap typical comparative
-analysis workflows. Moreover, by allowing the gathering, processing and analysis
-of data to be performed programmatically, \texttt{rotl} facilitates reproducible
-research practices.
+analysis workflows, and extends the degree to which R supports reproducable
+research in phylogenetics and comparative biology.
 
@@
 R code) as well as per-function documentation. There is also an extensive
-test-suite that covers both the internal functions that \textt{rotl} uses to
+test-suite that covers both the internal functions that \texttt{rotl} uses to
 connect to OTL and public functions users apply to acess and process data.
@@
 open data and broaden the adoption of open science practices in
-general. \texttt{rotl} contributes to this initiative, and data from OTL can be
-combined with other sources easily. For instance in Figure~\ref{fig:cat_map}, we
-show how we can obtain a map of the occurrences for some of the cat species
+general. For instance, the rOpenSci-developed pacakge \texttt{rTreeBase} allows
+users to acess phylogenies stored in treebase. \texttt{rotl} contributes to this
+initiative, and greatly extends the number of taxa for which phylogenetic data
+can be retreived within R while allowing the  data from OTL can be combined with 
+other sources easily.
+
+The package vignettes provide two demonstrations of the integration of a
+phylogeny and data associated the taxa it represents. Specifically, the 
+``Data mashups'' vignette provides an example of how data associated with tip-taxa
+can be gathered and visualized. Another vigenette titled ``meta-analysis'' 
+demonstrates how a complete analysis, including the gathering of data and a
+phylogeny, can be performed in a single R session. We further demonstrate the
+integration of phylogenetic data retrieved with \texttt{rotl} with other other
+data here. We how how we can obtain a map of the occurrences for some of the cat species
 (genus \textit{Lynx}) included in the phylogeny above using records for these
-species found in GBIF (code available in Appendix). Another example is available
-in the package vignette ``Data mashups''.
+species found in GBIF (code available in Appendix)
 
@@
 by the National Science Foundation’s Advancing Digitization of Biodiversity
-Collections Program (Cooperative Agreement EF-1115210)
+Collections Program (Cooperative Agreement EF-1115210).
 

### 2015-10-26 manuscript.Rnw
manuscript edits: now with spelling\!

@@
 associated with taxa in a given phylogeny can be obtained directly in R. 
-These packages, combined the language's support for literatre programming 
-\hl{(Knuth\_1984, Xie\_2014)} make R an ideal platform for reproducable research
-in phylogentics and comparative biology, as they allow a complete record of the 
+These packages, combined the language's support for literature programming 
+\hl{(Knuth\_1984, Xie\_2014)} make R an ideal platform for reproducible research
+in phylogenetics and comparative biology, as they allow a complete record of the 
 steps taken in gathering, processing and analysing a given data set to be
@@
 Here we present \texttt{rotl}, and R package that allows users to download
-phylogentic and taxonomic data from Open Tree of Life in R. \texttt{rotl} takes
-adavantage of Application Programming Interfaces (APIs) provided by OTL to acess
+phylogenetic and taxonomic data from Open Tree of Life in R. \texttt{rotl} takes
+advantage of Application Programming Interfaces (APIs) provided by OTL to acess
 sub-trees from the synthetic Open Tree and the published trees that contribute
@@
 general. For instance, the rOpenSci-developed pacakge \texttt{rTreeBase} allows
-users to acess phylogenies stored in treebase. \texttt{rotl} contributes to this
+users to access phylogenies stored in treebase. \texttt{rotl} contributes to this
 initiative, and greatly extends the number of taxa for which phylogenetic data
-can be retreived within R while allowing the  data from OTL can be combined with 
+can be retrieved within R while allowing the  data from OTL can be combined with 
 other sources easily.
@@
 ``Data mashups'' vignette provides an example of how data associated with tip-taxa
-can be gathered and visualized. Another vigenette titled ``meta-analysis'' 
+can be gathered and visualized. Another vignette titled ``meta-analysis'' 
 demonstrates how a complete analysis, including the gathering of data and a

### 2015-10-28 manuscript.Rnw
jwb edits

--- 2015-10-21 manuscript.Rnw
@@
 While phylogenies have been getting easier to build, it has been difficult to
-re-use, combine and synthesize the information they provide because published
+re-use, combine, and synthesize the information they provide because published
 trees are often only available as image files, and taxonomic information is not
@@
 provides tools and services to query and download parts of this synthetic tree,
-as well as the data used to build it. Here, we present \texttt{rotl}, an R
-package to search and download data from the Open Tree of Life data directly in
+as well as the source data used to build it. Here, we present \texttt{rotl}, an R
+package to search and download data from the Open Tree of Life directly in
 R. \texttt{rotl} uses common data structures allowing researchers to take
 advantage of the rich set of tools and methods that are available in R to
-manipulate, analyze and visualize phylogenies.
+manipulate, analyze, and visualize phylogenies.
 
@@
 disputes and estimate phylogenies for previously understudied groups. In
-parallel, the ease which which phylogenies can be estimated has lead to the
-development of new phylogenetic comparative methods. These methods allow
-researchers to explore fundamental questions about the origin of biodiversity
+parallel, the ease with which phylogenies can be estimated has spurred the
+development of novel phylogenetic comparative methods. These methods allow
+researchers to explore fundamental questions about the origin of biodiversity,
 including the evolution of morphological and ecological traits, the
@@
 
-Ideally, the ever increasing number of published phylogenies would contribute to
-a synthesis of phylogenetic knowledge -- leading to a better understanding of
-the history of life and providing high-quality phylogenetic information for use
-in comparative methods. However, in practice synthesizing phylogenetic data
-remains difficult. Phylogenies are scattered, often only available as image
+Ideally, the ever-increasing number of published phylogenies would contribute to
+a synthesis of phylogenetic knowledge, ultimately leading to a richer understanding of
+the history of life, while at the same time providing high-quality phylogenetic information for use
+in comparative analyses. However, in practice synthesizing phylogenetic data
+is a difficult task. Phylogenetic information is scattered, often only available as image
 files within publications, and the lack of standardization to store and
@@
 synthesize, and integrate this information into their own research
-(\citealt{Stoltzfus2012}, but see \citealt{Cranston2014} for suggestions of best
+(\citealt{Stoltzfus2012}, \citealt{Drew2013}; but see \citealt{Cranston2014} for suggestions of best
 practices).
@@
 elucidate relationships among taxa. This combination of information is used to
-form the synthetic tree. Studies can be contributed to the synthetic tree
-through the curator interface \url{https://tree.opentreeoflife.org/curator},
+form a comprehensive synthetic tree. Studies can be contributed to the synthetic tree
+through a curator interface (\url{https://tree.opentreeoflife.org/curator}),
 allowing the synthetic tree to be continuously updated as relationships are
-elucidated or reevaluated. The current draft of the Open Tree of Life contains
+elucidated or reevaluated. The current draft of the OTL synthetic tree contains
 2.3 million tips. Beyond obvious applications across the life sciences to
 explore questions in evolution, biodiversity, and conservation, the resources
-OTL provides could be used for education and outreach (e.g., illustrating course
-material, develop outreach activities to explore relationships among species).
+OTL provides is useful for education and outreach (e.g., illustrating course
+material, or developing outreach activities to explore relationships among species).
 
@@
 
-The Open Tree of Life project provides a web interface to explore, query and
-download parts of the synthetic Open Tree (\url{http://tree.opentreeoflife.org})
-and the published trees that contribute to it. Additionally, the project
+The OTL project provides a web interface to explore, query and
+download parts of the synthetic tree (\url{http://tree.opentreeoflife.org})
+as well as the published trees that contribute to it. Additionally, the project
 provides Application Programming Interfaces (APIs)  to access programmatically
-the different services and underlying data.  Here, we present \texttt{rotl} a
+the different services and underlying data.  Here, we present \texttt{rotl}, a
 package to interact with OTL's APIs to search for and import phylogenetic trees
-and taxonomic information directly in R. Having access to this data in R allows
+and taxonomic information directly in R. Having access to these data in R allows
 researchers to take advantage of the multiple packages that are available to
-visualize, analyze and manipulate phylogenetic and comparative data (e.g., ape
+visualize, analyze, and manipulate phylogenetic and comparative data (e.g., ape
 \citealt{Paradis2004}, phytools \citeateu{Revell2012}, geiger
 \citealt{Pennell2014-geiger}, ggtree \citealt{Yu-ggtree}, phylobase
-\citealt{phylobase-0.8.0}, RNeXML \citealt{Boettiger2015}, see
+\citealt{phylobase-0.8.0}, RNeXML \citealt{Boettiger2015}; see
 \url{https://cran.r-project.org/web/views/Phylogenetics.html} for a
 comprehensive list). There is also a growing list of packages that allow users
-to query and acces data from the web directly in R \hl{rFISHBASE
+to query and access data from the web directly in R: \hl{rFISHBASE
 10.1111/j.1095-8649.2012.03464.x, paleoDB 10.1111/ecog.01154, rAvis
 10.1371/journal.pone.0091650 \dots}. By providing direct access to high quality
-phylogenetic data in R, \texttt{rotl} fills a key gap typical comparative
-analysis workflows. Moreover, by allowing the gathering, processing and analysis
+phylogenetic data in R, \texttt{rotl} fills a key gap in typical comparative
+analysis workflows. Moreover, by allowing the gathering, processing, and analysis
 of data to be performed programmatically, \texttt{rotl} facilitates reproducible
@@
 \begin{enumerate}
-\item The \emph{taxonomy} used as the backbone from the tree, the Open Tree
-  Taxonomy;
-\item The \emph{studies} and their associated trees some of which are chosen by
+\item The \emph{taxonomy} used as the backbone of the tree, the Open Tree
+  Taxonomy (OTT);
+\item The \emph{studies} and their associated trees, some of which are chosen by
   curators to assemble the synthetic tree;
@@
   names to the Open Tree Taxonomy identifiers;
-\item The \emph{synthetic tree} itself, the Open Tree.
+\item The \emph{synthetic tree} itself, the `Open Tree'.
 \end{enumerate}
@@
 used to synthesize the various sources of information from which the Open Tree
-is assembled from. Data from this API is not intended for normal users, and its
-use will not be demonstrated in this paper but \texttt{rotl} has functions to
+is assembled. Data from this API are not intended for normal users, and its
+use will not be demonstrated in this paper, although \texttt{rotl} has functions to
 retrieve data it provides.
@@
 \texttt{rotl} gives users access to all endpoints provided by version 2 of the
-OTL's APIs. Phylogenetic trees served by the API can be imported directly into
+OTL APIs. Phylogenetic trees served by the API can be imported directly into
 R's memory and are represented using the \texttt{ape} \citep{Paradis2004} tree
@@
 allows researchers to use these trees either directly with other R packages, or
-to be imported in other programs that can read in phylogenetic tree files.
+to be imported in other programs that make use of phylogenetic tree files.
 
-The synthetic tree currently does not have any branch lengths associated with
+The full Open Tree currently does not have any branch lengths associated with
 it, therefore parametric comparative methods cannot be used directly on the
-trees returned by OTL. However, resources and methods are being developed to add
-branch lengths to these topological trees \citep[e.g.,][]{Ksepka2015} or use
+subtrees returned by OTL (although the OTL treestore contains the raw 
+published source phylogenies, complete with branch lengths and node 
+annotations; see below). However, resources and methods are being developed to add
+branch lengths to these topological subtrees \citep[e.g.,][]{Ksepka2015} or use
 topological trees to identify phylogenetically equivalent species to increase
 overlap between chronograms and species trait data \citep{Pennell2015}. Without
-branch lengths, these trees are nonetheless useful to illustrate relationships
+branch lengths, these subtrees are nonetheless useful to illustrate relationships
 among species, or to map a trait on a phylogeny for instance.
@@
 \citep[NCL,][]{Lewis2003} as implemented in the \texttt{rncl} package
-\url{https://cran.r-project.org/package=rncl}. Using NCL provides robust and
+(\url{https://cran.r-project.org/package=rncl}). Using NCL provides robust and
 efficient parsing of large trees that may contain singleton nodes labeled with
@@
 vignettes (documents that demonstrate the use of the package and contain executable 
-R code) as well as per-function documentation. There is also an extensive
-test-suite that covers both the internal functions that \textt{rotl} uses to
-connect to OTL and public functions users apply to acess and process data.
+R code)
+% as well as per-function documentation
+. There is also an extensive
+test-suite that covers both the internal functions that \texttt{rotl} uses to
+connect to OTL and the public functions users apply to access and process data.
 
@@
 To get the relationships among a set of taxa from the Open Tree, the taxa first
-need to be matched against the Open Taxonomy using the TNRS. This step retrieves
+need to be matched against the OTL taxonomy OTT using the TNRS. This step retrieves
 the identifiers that will be used to extract the relationships for the set of
@@
 OTL to draw a phylogenetic tree for a set of model organisms.
-
+First, we use the function \texttt{tnrs\_match\_names} to match the taxon 
+names to their Open Taxonomy identifiers.
 <<init, echo=FALSE, purl=FALSE>>=
@@
 
-First, we use the function \texttt{tnrs\_match\_names} to match the taxon names to
-their Open Taxonomy identifiers.
+%First, we use the function \texttt{tnrs\_match\_names} to match the taxon names to their Open Taxonomy identifiers.
 
@@
 taxa matched are the correct ones. Here, there is no ambiguity in the taxa
-matched, however, as the Open Tree Taxonomy includes taxa from bacteria, plants,
-and animals that are regulated by different nomenclatural codes (ICNP, ICN and
-ICZN respectively), OTL and \texttt{rotl} provide tools to deal with potential
+matched; however, as OTT includes taxa from bacteria, plants,
+and animals that are regulated by different nomenclatural codes (ICNP, ICN, and
+ICZN, respectively), both OTL and \texttt{rotl} provide tools to deal with potential
 hemihomonyms. The argument \texttt{context\_name} can be used to limit potential
 matches to a taxonomic group such as ``Animals'' (see the function
-\texttt{tnrs\_contexts} for a list of possible options).  When this strategy
-cannot be used in case the tree encompasses multiple domains (as in the present
-example), the function \texttt{inspect} lists alternative matches for a taxon
+\texttt{tnrs\_contexts} for a complete list of possible options).  When this strategy
+cannot be used (as in the present example, where the tree encompasses 
+multiple domains), the function \texttt{inspect} lists alternative matches for a taxon
 name, and \texttt{update} replaces it in the results. An example of this
@@
 By default, approximate matching is enabled when attempting to match taxonomic
-names to their Open Tree Taxonomy identifiers. Additionally, taxonomic synonyms
-are included in the Open Tree Taxonomy allowing researchers to match correct
+names to their OTT identifiers. Additionally, taxonomic synonyms
+are included in OTT, allowing researchers to match correct
 identifiers for taxon names that might include misspellings or synonyms. These
 features will facilitate the tedious data cleaning process often needed when
-matching taxon names. In the example provided, both \textit{Escherichia coli}
-and \textit{Saccharomyces cerevisiae} are misspelled but OTL's TNRS finds the
+mapping taxon names. In the example provided, both \textit{Escherichia coli}
+and \textit{Saccharomyces cerevisiae} are misspelled, but the OTL TNRS finds the
 correct match for these taxa.
@@
 contribute to the synthetic tree. These trees provide a useful resource to
-reproduce or expand on an already published analysis, or to explore how the
+reproduce or expand on a previously published analysis, or to explore how the
 elucidation of relationships within a clade has changed through time.
@@
 \texttt{studies\_properties}. Typically, users will want to search for studies
-or trees based on taxon names (or their Open Taxonomy identifiers), but other
+or trees based on taxon names (or their OTT identifiers), but other
 criteria such as the title of the publication can be used. Here we demonstrate
@@
 
-In recent years, R has become part of the toolbox of many researchers in
+In recent years, R has become an essential part of the toolbox of many researchers in
 evolutionary biology. R greatly facilitates the analysis of large datasets, and
@@
 comparative analyses are implemented, and because it is a relatively easy to use
-programming language. Additionally, as more data is made available online and
+programming language. Additionally, as more data are made available online and
 accessible using APIs, several packages have been developed to interact and
-download these datasets directly in R enabling direct and reproducible analyses.
-Notably, the organization rOpenSci \url{https://ropensci.org} has fostered a
+download these datasets directly in R, enabling direct and reproducible analyses.
+Notably, the organization rOpenSci (\url{https://ropensci.org}) has fostered a
 community of researchers who develop tools and methods to facilitate the use of
-open data and broaden the adoption of open science practices in
+open data, as well as broaden the adoption of open science practices in
 general. \texttt{rotl} contributes to this initiative, and data from OTL can be
@@
 (genus \textit{Lynx}) included in the phylogeny above using records for these
-species found in GBIF (code available in Appendix). Another example is available
+species found in GBIF (code available in the Appendix). Another example is available
 in the package vignette ``Data mashups''.
@@
 ## load required packages
+library(rotl)
 library(rgbif)
@@
 library(maps)
+
+## Import the felid tree using study and tree IDs discovered with 
+## studies_find_studies() in the manuscript
+
+cat_tree <- get_study_tree(study_id ="pg_1981",                           
+                           tree_id = "tree4052")
 
@@
 non-independence of species in comparative methods, and the recent development
-of methods to explore trait evolution or changes in diversification rates have
+of methods to explore trait evolution or changes in diversification rates, have
 driven the need for accurate phylogenies. However, there is often a discrepancy
@@
 submit their published phylogenies to OTL using the curator interface
-\url{https://tree.opentreeoflife.org/curator}. By facilitating the discovery and
-re-use of published trees and of the synthetic tree, we hope \texttt{rotl} will
+(\url{https://tree.opentreeoflife.org/curator}). By facilitating the discovery and
+re-use of published trees and of the synthetic Open Tree, we hope \texttt{rotl} will
 contribute to the wider adoption of best practices to make phylogenetic
@@
 BSD license. Stable versions are available from the CRAN repository
-\url{https://cran.r-project.org/package=rotl}, and development
+(\url{https://cran.r-project.org/package=rotl}), and development
 versions are available from GitHub
-\url{https://github.com/ropensci/rotl}. The package is under active development,
+(\url{https://github.com/ropensci/rotl}). The package is under active development,
 and authors welcome bug reports of feature requests via the github repository.
@@
 
-Python \url{https://github.com/OpenTreeOfLife/pyopentree} and Ruby
-\url{https://github.com/SpeciesFileGroup/bark} libraries to interact
+Python (\url{https://github.com/OpenTreeOfLife/pyopentree}) and Ruby
+(\url{https://github.com/SpeciesFileGroup/bark}) libraries to interact
 with the OTL APIs are also available.
@@
 We would like to thank the organizers of the OpenTree of Life APIs hackathon
-that was held at the University of Michigan, Ann Arbor, September 15-19, 2014,
+that was held at the University of Michigan, Ann Arbor, 15-19 September, 2014,
 where the development of \texttt{rotl} was started. We would also like to thank

### 2015-10-28 manuscript.Rnw
appendix

--- 2015-10-26 manuscript.Rnw
@@
 library(maps)
+
+## Import the felid tree using study and tree IDs discovered with 
+## studies_find_studies() in the manuscript
+cat_tree <- get_study_tree(study_id ="pg_1981",                           
+                           tree_id = "tree4052")
 

### 2015-10-29 manuscript.Rnw
jay-double-ewe-bee edits

--- 2015-10-28 manuscript.Rnw
@@
 parallel, the ease with which phylogenies can be estimated has spurred the
-development of novel phylogenetic comparative methods. These methods allow
-researchers to explore fundamental questions about the origin of biodiversity,
+development of new phylogenetic comparative methods. These methods allow
+researchers to explore fundamental questions about the origin of biodiversity
 including the evolution of morphological and ecological traits, the
 spatio-temporal variation in speciation rates, or both
-\citep{OMeara2012,Pennell2013}.
+\citep{OMeara2012, Pennell2013}.
 
-Ideally, the ever-increasing number of published phylogenies would contribute to
-a synthesis of phylogenetic knowledge, ultimately leading to a richer understanding of
-the history of life, while at the same time providing high-quality phylogenetic information for use
-in comparative analyses. However, in practice synthesizing phylogenetic data
-is a difficult task. Phylogenetic information is scattered, often only available as image
+Ideally, the ever increasing number of published phylogenies would contribute to
+a synthesis of phylogenetic knowledge, ultimately leading to a better understanding of
+the history of life while at the same time providing high-quality phylogenetic information 
+for use in comparative analyses. However, in practice synthesizing phylogenetic data
+is a difficult task. Phylogenetic information is largely scattered, often only available as image
 files within publications, and the lack of standardization to store and
@@
 synthesize, and integrate this information into their own research
-(\citealt{Stoltzfus2012}, \citealt{Drew2013}; but see \citealt{Cranston2014} for suggestions of best
+(\citealt{Stoltzfus2012}; \citealt{Drew2013}; but see \citealt{Cranston2014} for suggestions of best
 practices).
@@
 elucidate relationships among taxa. This combination of information is used to
-form a comprehensive synthetic tree. Studies can be contributed to the synthetic tree
-through a curator interface (\url{https://tree.opentreeoflife.org/curator}),
+structure the comprehensive synthetic tree. Studies can be contributed to the synthetic 
+tree through a curator interface (\url{https://tree.opentreeoflife.org/curator}),
 allowing the synthetic tree to be continuously updated as relationships are
@@
 
-The OTL project provides a web interface to explore, query and
-download parts of the synthetic tree (\url{http://tree.opentreeoflife.org})
-as well as the published trees that contribute to it. Additionally, the project
-provides Application Programming Interfaces (APIs)  to access programmatically
-the different services and underlying data.  Here, we present \texttt{rotl}, a
-package to interact with OTL's APIs to search for and import phylogenetic trees
-and taxonomic information directly in R. Having access to these data in R allows
-researchers to take advantage of the multiple packages that are available to
-visualize, analyze, and manipulate phylogenetic and comparative data (e.g., ape
-\citealt{Paradis2004}, phytools \citeateu{Revell2012}, geiger
-\citealt{Pennell2014-geiger}, ggtree \citealt{Yu-ggtree}, phylobase
-\citealt{phylobase-0.8.0}, RNeXML \citealt{Boettiger2015}; see
+The R programming language is a popular tool for  phylogenetics and comparative
+analysis. The R packages \texttt{ape} \citep{Paradis2004}, \texttt{phylobase}
+\citep{phylobase-0.8.0}, and \texttt{RNeXML} \citep{Boettiger2015} each provide
+functions to import and manipulate trees within R and save the results in
+standard data formats. Additional packages including \texttt{phytools}
+\citep{Revell2012}, \texttt{geiger} \citep{Pennell2014-geiger}, and \texttt{ggtree} 
+\citep{Yu-ggtree} allow users to analyse and visualise data in a 
+phylogenetic context (see 
 \url{https://cran.r-project.org/web/views/Phylogenetics.html} for a
-comprehensive list). There is also a growing list of packages that allow users
-to query and access data from the web directly in R: \hl{rFISHBASE
-10.1111/j.1095-8649.2012.03464.x, paleoDB 10.1111/ecog.01154, rAvis
-10.1371/journal.pone.0091650 \dots}. By providing direct access to high quality
-phylogenetic data in R, \texttt{rotl} fills a key gap in typical comparative
-analysis workflows. Moreover, by allowing the gathering, processing, and analysis
-of data to be performed programmatically, \texttt{rotl} facilitates reproducible
-research practices.
+comprehensive list of phylogenetics packages in R). In addition to packages for
+phylogentic and comparative analyses, a growing number of R packages allow users to query and
+access data from the web (e.g. \texttt{rFISHBASE} \citep{Boettiger2012}, \texttt{rAvis} \citep{Varela2014}, and \texttt{paleoDB} \citep{Varela2015}), meaning data 
+associated with taxa in a given phylogeny can be obtained directly in R. 
+These packages, combined with the language's support for literate programming 
+(\citealt{Knuth1984}; \citealt{Xie2014}), make R an ideal platform for reproducible research
+in phylogenetics and comparative biology, as they allow a complete record of the 
+steps taken in gathering, processing, and analysing a given data set to be
+produced.
+
+Here we present \texttt{rotl}, and R package that allows users to download
+phylogenetic and taxonomic data from the OTL directly in R. \texttt{rotl} takes
+advantage of Application Programming Interfaces (APIs) provided by OTL to access
+sub-trees from the synthetic Open Tree, as well as the published source trees that 
+contribute to the synthesis. By providing direct access to high quality
+phylogenetic data in R, \texttt{rotl} fills a key gap typical comparative
+analysis workflows, and extends the degree to which R supports reproducable
+research in phylogenetics and comparative biology.
 
@@
   Taxonomy (OTT);
-\item The \emph{studies} and their associated trees, some of which are chosen by
+\item The \emph{studies} and their associated trees some of which are chosen by
   curators to assemble the synthetic tree;
@@
 Additionally, the project provides access to other data, the Graph of Life, that
-stores information about the underlying tree alignment graphs \citep{Smith2013}
+stores information about the underlying tree alignment graph \citep{Smith2013}
 used to synthesize the various sources of information from which the Open Tree
@@
 structure (objects of class \texttt{phylo}), or can be written to files in the
-Newick, NEXUS \citep{Maddison1997} or NeXML \citep{Vos2012} file formats. This
+Newick, NEXUS \citep{Maddison1997}, or NeXML \citep{Vos2012} file formats. This
 allows researchers to use these trees either directly with other R packages, or
@@
 
-% Tree versionning? Replicability of analyses
+% Tree versioning? Replicability of analyses
 
@@
 R code)
-% as well as per-function documentation
+%as well as per-function documentation
 . There is also an extensive
 test-suite that covers both the internal functions that \texttt{rotl} uses to
-connect to OTL and the public functions users apply to access and process data.
+connect to OTL and public functions users apply to acess and process data.
 
@@
 To illustrate how to obtain relationships from a set of taxa, we will rely on
-OTL to draw a phylogenetic tree for a set of model organisms.
-First, we use the function \texttt{tnrs\_match\_names} to match the taxon 
-names to their Open Taxonomy identifiers.
+OTL to draw a phylogenetic tree for a set of model organisms. First, we use the function 
+\texttt{tnrs\_match\_names} to match the taxon names to their Open Taxonomy identifiers.
+
 <<init, echo=FALSE, purl=FALSE>>=
@@
 @
-
-%First, we use the function \texttt{tnrs\_match\_names} to match the taxon names to their Open Taxonomy identifiers.
-
 <<get_taxa, cache=TRUE, purl=FALSE>>=
@@
 tree <- tol_induced_subtree(ott_ids = taxa[["ott_id"]])
-plot(tree, cex=.8, label.offset=.1, no.margin=TRUE)
+plot(tree, cex = .8, label.offset = .1, no.margin = TRUE)
 @
@@
 
-<<plot_cats, purl=FALSE, fig.height=4.2, echo=FALSE, fig.cap='Phylogeny of the Felidae published in \\citealt{Johnson2006} and retrieved from OTL using \\texttt{rotl}'>>=
+<<plot_cats, purl=FALSE, fig.height=4.2, echo=FALSE, fig.cap='Phylogeny of the Felidae published in \\citealt{Johnson2006} and retrieved from OTL using \\texttt{rotl}.'>>=
 plot(cat_tree, label.offset=.0001, cex=.75, no.margin=TRUE)
@@
 accessible using APIs, several packages have been developed to interact and
-download these datasets directly in R, enabling direct and reproducible analyses.
+download these datasets directly in R, thereby enabling direct and reproducible analyses.
 Notably, the organization rOpenSci (\url{https://ropensci.org}) has fostered a
 community of researchers who develop tools and methods to facilitate the use of
-open data, as well as broaden the adoption of open science practices in
-general. \texttt{rotl} contributes to this initiative, and data from OTL can be
-combined with other sources easily. For instance in Figure~\ref{fig:cat_map}, we
-show how we can obtain a map of the occurrences for some of the cat species
+open data as well broaden the adoption of open science practices in
+general. For instance, the rOpenSci-developed packge \texttt{TreeBase} \citep{Boettiger2012TB} allows
+users to access phylogenies stored in treeBASE (\url{www.treebase.org}). 
+\texttt{rotl} contributes to this
+initiative, and greatly extends the number of taxa for which phylogenetic data
+can be retrieved within R, while allowing the  data from OTL can be combined with 
+other sources easily.
+
+The package vignettes provide two demonstrations of the integration of a
+phylogeny and data associated the taxa it represents. Specifically, the 
+``Data mashups'' vignette provides an example of how data associated with tip-taxa
+can be gathered and visualized. Another vignette titled ``meta-analysis'' 
+demonstrates how a complete analysis, including the gathering of data and a
+phylogeny, can be performed in a single R session. We further demonstrate the
+integration of phylogenetic data retrieved using \texttt{rotl} with other 
+data here. We show how we can obtain a map of the occurrences for some of the cat species
 (genus \textit{Lynx}) included in the phylogeny above using records for these
-species found in GBIF (code available in the Appendix). Another example is available
-in the package vignette ``Data mashups''.
+%species found in GBIF (Figure~\ref{fig:cat_map}; code available in the Appendix).
 
-<<cat_map, purl=TRUE, echo=FALSE, size='small', fig.height=4.2, results='hide', message=FALSE, cache=TRUE, fig.cap='GBIF records for the species in \\textit{Lynx} included in the phylogeny associated with the study by \\citealt{Johnson2006}'>>=
+<<cat_map, purl=TRUE, echo=FALSE, size='small', fig.height=4.2, results='hide', message=FALSE, cache=TRUE, fig.cap='GBIF records for the species in \\textit{Lynx} included in the phylogeny associated with the study by \\citealt{Johnson2006}.'>>=
 ## load required packages
@@
 ## studies_find_studies() in the manuscript
-
 cat_tree <- get_study_tree(study_id ="pg_1981",                           
@@
 and authors welcome bug reports of feature requests via the github repository.
-The source for this manuscript is available at \url{https://github.com/fmichonneau/rotl-ms}.
+The source for this manuscript is available on GitHub (\url{https://github.com/fmichonneau/rotl-ms}).
 

### 2015-10-29 manuscript.Rnw
some spelling

@@
 \citep{Revell2012}, \texttt{geiger} \citep{Pennell2014-geiger}, and \texttt{ggtree} 
-\citep{Yu-ggtree} allow users to analyse and visualise data in a 
+\citep{Yu-ggtree} allow users to analyze and visualize data in a 
 phylogenetic context (see 
@@
 comprehensive list of phylogenetics packages in R). In addition to packages for
-phylogentic and comparative analyses, a growing number of R packages allow users to query and
+phylogenetic and comparative analyses, a growing number of R packages allow users to query and
 access data from the web (e.g. \texttt{rFISHBASE} \citep{Boettiger2012}, \texttt{rAvis} \citep{Varela2014}, and \texttt{paleoDB} \citep{Varela2015}), meaning data 
@@
 in phylogenetics and comparative biology, as they allow a complete record of the 
-steps taken in gathering, processing, and analysing a given data set to be
+steps taken in gathering, processing, and analyzing a given data set to be
 produced.
@@
 phylogenetic data in R, \texttt{rotl} fills a key gap typical comparative
-analysis workflows, and extends the degree to which R supports reproducable
+analysis workflows, and extends the degree to which R supports reproducible
 research in phylogenetics and comparative biology.
@@
 test-suite that covers both the internal functions that \texttt{rotl} uses to
-connect to OTL and public functions users apply to acess and process data.
+connect to OTL and public functions users apply to access and process data.
 

### 2015-11-02 manuscript.Rnw
[fix #6] improve figure

@@
     geom_point(aes(y = decimalLatitude, x = decimalLongitude, color = Species),
-               alpha = 0.3, size = 2) +
+               size = 1) +
     coord_map(projection = "mercator", orientation = c(90, 0, 0)) +
     xlab("Longitude") + ylab("Latitude") +
-    theme(legend.position="top") + ylim(c(0,72))
-
+    theme(legend.position="top",
+          legend.key = element_rect(fill = "gray40")) +
+    ylim(c(0,72))
 @

### 2015-11-02 manuscript.Rnw
fix typo/formatting

@@
 standardized across studies. The Open Tree of Life (OTL) project addresses these
-issues by providing a digital tree that encompasses all organisms built by
+issues by providing a digital tree that encompasses all organisms, built by
 combining taxonomic information and published phylogenies. The project also
 provides tools and services to query and download parts of this synthetic tree,
-as well as the source data used to build it. Here, we present \texttt{rotl}, an R
-package to search and download data from the Open Tree of Life directly in
+as well as the source data used to build it. Here, we present \texttt{rotl}, an
+R package to search and download data from the Open Tree of Life directly in
 R. \texttt{rotl} uses common data structures allowing researchers to take
@@
 Ideally, the ever increasing number of published phylogenies would contribute to
-a synthesis of phylogenetic knowledge, ultimately leading to a better understanding of
-the history of life while at the same time providing high-quality phylogenetic information 
-for use in comparative analyses. However, in practice synthesizing phylogenetic data
-is a difficult task. Phylogenetic information is largely scattered, often only available as image
-files within publications, and the lack of standardization to store and
-represent phylogenetic data makes it difficult for researchers to access,
-synthesize, and integrate this information into their own research
-(\citealt{Stoltzfus2012}; \citealt{Drew2013}; but see \citealt{Cranston2014} for suggestions of best
+a synthesis of phylogenetic knowledge, ultimately leading to a better
+understanding of the history of life while at the same time providing
+high-quality phylogenetic information for use in comparative analyses. However,
+in practice, synthesizing phylogenetic data is a difficult task. Phylogenetic
+information is largely scattered, often only available as image files within
+publications, and the lack of standardization to store and represent
+phylogenetic data makes it difficult for researchers to access, synthesize, and
+integrate this information into their own research (\citealt{Stoltzfus2012};
+\citealt{Drew2013}; but see \citealt{Cranston2014} for suggestions of best
 practices).
@@
 elucidate relationships among taxa. This combination of information is used to
-structure the comprehensive synthetic tree. Studies can be contributed to the synthetic 
-tree through a curator interface (\url{https://tree.opentreeoflife.org/curator}),
-allowing the synthetic tree to be continuously updated as relationships are
-elucidated or reevaluated. The current draft of the OTL synthetic tree contains
-2.3 million tips. Beyond obvious applications across the life sciences to
-explore questions in evolution, biodiversity, and conservation, the resources
-OTL provides is useful for education and outreach (e.g., illustrating course
-material, or developing outreach activities to explore relationships among species).
+structure the comprehensive synthetic tree. Studies can be contributed to the
+synthetic tree through a curator interface
+(\url{https://tree.opentreeoflife.org/curator}), allowing the synthetic tree to
+be continuously updated as relationships are elucidated or reevaluated. The
+current draft of the OTL synthetic tree contains 2.3 million tips. Beyond
+obvious applications across the life sciences to explore questions in evolution,
+biodiversity, and conservation, the resources OTL provides are useful for
+education and outreach (e.g., illustrating course material, or developing
+outreach activities to explore relationships among species).
 
@@
 standard data formats. Additional packages including \texttt{phytools}
-\citep{Revell2012}, \texttt{geiger} \citep{Pennell2014-geiger}, and \texttt{ggtree} 
-\citep{Yu-ggtree} allow users to analyze and visualize data in a 
-phylogenetic context (see 
+\citep{Revell2012}, \texttt{geiger} \citep{Pennell2014-geiger}, and
+\texttt{ggtree} \citep{Yu-ggtree}, allow users to analyze and visualize data in
+a phylogenetic context (see
 \url{https://cran.r-project.org/web/views/Phylogenetics.html} for a
 comprehensive list of phylogenetics packages in R). In addition to packages for
-phylogenetic and comparative analyses, a growing number of R packages allow users to query and
-access data from the web (e.g. \texttt{rFISHBASE} \citep{Boettiger2012}, \texttt{rAvis} \citep{Varela2014}, and \texttt{paleoDB} \citep{Varela2015}), meaning data 
-associated with taxa in a given phylogeny can be obtained directly in R. 
-These packages, combined with the language's support for literate programming 
-(\citealt{Knuth1984}; \citealt{Xie2014}), make R an ideal platform for reproducible research
-in phylogenetics and comparative biology, as they allow a complete record of the 
-steps taken in gathering, processing, and analyzing a given data set to be
-produced.
+phylogenetic and comparative analyses, a growing number of R packages allows
+users to query and access data from the web (e.g. \texttt{rFISHBASE}
+\citep{Boettiger2012}, \texttt{rAvis} \citep{Varela2014}, and \texttt{paleoDB}
+\citep{Varela2015}), such that data associated with taxa in a given phylogeny
+can be obtained directly in R.  These packages, combined with the language's
+support for literate programming (\citealt{Knuth1984}; \citealt{Xie2014}), make
+R an ideal platform for reproducible research in phylogenetics and comparative
+biology, as they allow a complete record of the steps taken in gathering,
+processing, and analyzing a given data set to be produced.
 
@@
 phylogenetic and taxonomic data from the OTL directly in R. \texttt{rotl} takes
-advantage of Application Programming Interfaces (APIs) provided by OTL to access
-sub-trees from the synthetic Open Tree, as well as the published source trees that 
+advantage of OTL's Application Programming Interfaces (APIs) to access sub-trees
+from the synthetic Open Tree, as well as the published source trees that
 contribute to the synthesis. By providing direct access to high quality
-phylogenetic data in R, \texttt{rotl} fills a key gap typical comparative
+phylogenetic data in R, \texttt{rotl} fills a key gap in typical comparative
 analysis workflows, and extends the degree to which R supports reproducible
@@
 To get the relationships among a set of taxa from the Open Tree, the taxa first
-need to be matched against the OTL taxonomy OTT using the TNRS. This step retrieves
-the identifiers that will be used to extract the relationships for the set of
-requested taxa.
+needs to be matched against the OTL taxonomy (OTT) using the TNRS. This step
+retrieves the identifiers that will be used to extract the relationships for the
+set of requested taxa.
 
@@
 been submitted through the curator interface, and identify the trees that
-contribute to the synthetic tree. These trees provide a useful resource to
+contribute to the synthetic tree. These trees constitute a useful resource to
 reproduce or expand on a previously published analysis, or to explore how the
@@
 
-In recent years, R has become an essential part of the toolbox of many researchers in
-evolutionary biology. R greatly facilitates the analysis of large datasets, and
-allows researchers to combine methods in novel ways because many methods for
-comparative analyses are implemented, and because it is a relatively easy to use
-programming language. Additionally, as more data are made available online and
-accessible using APIs, several packages have been developed to interact and
-download these datasets directly in R, thereby enabling direct and reproducible analyses.
-Notably, the organization rOpenSci (\url{https://ropensci.org}) has fostered a
-community of researchers who develop tools and methods to facilitate the use of
-open data as well broaden the adoption of open science practices in
-general. For instance, the rOpenSci-developed packge \texttt{TreeBase} \citep{Boettiger2012TB} allows
-users to access phylogenies stored in treeBASE (\url{www.treebase.org}). 
-\texttt{rotl} contributes to this
-initiative, and greatly extends the number of taxa for which phylogenetic data
-can be retrieved within R, while allowing the  data from OTL can be combined with 
-other sources easily.
+In recent years, R has become an essential part of the toolbox of many
+researchers in evolutionary biology. R greatly facilitates the analysis of large
+datasets, and allows researchers to combine methods in novel ways because many
+methods for comparative analyses are implemented, and because it is a relatively
+easy to use programming language. Additionally, as more data are made available
+online and accessible using APIs, several packages have been developed to
+interact and download these datasets directly in R, thereby enabling direct and
+reproducible analyses.  Notably, the organization rOpenSci
+(\url{https://ropensci.org}) has fostered a community of researchers who develop
+tools and methods to facilitate the use of open data as well broaden the
+adoption of open science practices in general. For instance, the
+rOpenSci-developed packge \texttt{TreeBase} \citep{Boettiger2012TB} allows users
+to access phylogenies stored in treeBASE (\url{www.treebase.org}).
+\texttt{rotl} contributes to this initiative, and greatly extends the number of
+taxa for which phylogenetic data can be retrieved within R, while allowing the
+data from OTL can be combined with other sources easily.
 
 The package vignettes provide two demonstrations of the integration of a
-phylogeny and data associated the taxa it represents. Specifically, the 
-``Data mashups'' vignette provides an example of how data associated with tip-taxa
-can be gathered and visualized. Another vignette titled ``meta-analysis'' 
+phylogeny and data associated the taxa it represents. Specifically, the ``Data
+mashups'' vignette provides an example of how data associated with tip-taxa can
+be gathered and visualized. Another vignette titled ``meta-analysis''
 demonstrates how a complete analysis, including the gathering of data and a
 phylogeny, can be performed in a single R session. We further demonstrate the
-integration of phylogenetic data retrieved using \texttt{rotl} with other 
-data here. We show how we can obtain a map of the occurrences for some of the cat species
-(genus \textit{Lynx}) included in the phylogeny above using records for these
-%species found in GBIF (Figure~\ref{fig:cat_map}; code available in the Appendix).
+integration of phylogenetic data retrieved using \texttt{rotl} with other data
+here. We show how we can obtain a map of the occurrences for some of the cat
+species (genus \textit{Lynx}) included in the phylogeny above using records for
+these species found in GBIF (Figure~\ref{fig:cat_map}; code available in the
+Appendix).
 
@@
     xlab("Longitude") + ylab("Latitude") +
-    theme(legend.position="top",
-          legend.key = element_rect(fill = "gray40")) +
+    theme(legend.position="top", legend.key = element_rect(fill = "gray40")) +
     ylim(c(0,72))
@@
 
-\texttt{rotl} is free, open source, and released under a Simplified
-BSD license. Stable versions are available from the CRAN repository
-(\url{https://cran.r-project.org/package=rotl}), and development
-versions are available from GitHub
-(\url{https://github.com/ropensci/rotl}). The package is under active development,
-and authors welcome bug reports of feature requests via the github repository.
-The source for this manuscript is available on GitHub (\url{https://github.com/fmichonneau/rotl-ms}).
+\texttt{rotl} is free, open source, and released under a Simplified BSD
+license. Stable versions are available from the CRAN repository
+(\url{https://cran.r-project.org/package=rotl}), and development versions are
+available from GitHub (\url{https://github.com/ropensci/rotl}). The package is
+under active development, and authors welcome bug reports of feature requests
+via the GitHub repository.  The source for this manuscript is available on
+GitHub (\url{https://github.com/fmichonneau/rotl-ms}).
 

### 2015-11-02 manuscript.Rnw
Added helpers + DJW grant support to acknowldegements

@@
 where the development of \texttt{rotl} was started. We would also like to thank
-Scott Chamberlain (rOpenSci) for providing a thorough code review. FM was
+Scott Chamberlain (rOpenSci) for providing a thorough code review and 
+Shinichi Nakagawa and Alistair Senior for their help in developing the package's
+meta-analysis vignette. DJW was supported by  NIH Grant R01-GM101352. SFM was
 supported by iDigBio, and therefore this  material is based upon work supported

### 2015-11-02 manuscript.Rnw
last minute edits

--- 2015-11-02 manuscript.Rnw
@@
 systematic biology. The ability to routinely generate molecular datasets from
-any extant organism has allowed researchers to resolve long-running taxonomic
+any extant organism has allowed researchers to resolve long-standing taxonomic
 disputes and estimate phylogenies for previously understudied groups. In
@@
   Taxonomy (OTT);
-\item The \emph{studies} and their associated trees some of which are chosen by
+\item The \emph{studies} and their associated trees, some of which are chosen by
   curators to assemble the synthetic tree;
@@
 branch lengths, these subtrees are nonetheless useful to illustrate relationships
-among species, or to map a trait on a phylogeny for instance.
+among species, or to map traits on a phylogeny.
 
@@
 vignettes (documents that demonstrate the use of the package and contain executable
-R code)
-%as well as per-function documentation
-. There is also an extensive
-test-suite that covers both the internal functions that \texttt{rotl} uses to
-connect to OTL and public functions users apply to access and process data.
+R code). There is also an extensive test-suite that covers both the internal
+functions that \texttt{rotl} uses to connect to OTL and public functions users
+apply to access and process data.
 
@@
 To get the relationships among a set of taxa from the Open Tree, the taxa first
-needs to be matched against the OTL taxonomy (OTT) using the TNRS. This step
+need to be matched against the OTL taxonomy (OTT) using the TNRS. This step
 retrieves the identifiers that will be used to extract the relationships for the
@@
 criteria such as the title of the publication can be used. Here we demonstrate
-how to look for and retrieve a tree for studies focusing on the family Felidae
+how to look for and retrieve trees for studies focusing on the family Felidae
 (Figure~\ref{fig:plot_cats}).
@@
 taxa for which phylogenetic data can be retrieved within R, while allowing the
-data from OTL can be combined with other sources easily.
+data from OTL to be combined with other sources easily.
 
 The package vignettes provide two demonstrations of the integration of a
-phylogeny and data associated the taxa it represents. Specifically, the ``Data
+phylogeny and data associated with the taxa it represents. Specifically, the ``Data
 mashups'' vignette provides an example of how data associated with tip-taxa can
@@
 available from GitHub (\url{https://github.com/ropensci/rotl}). The package is
-under active development, and authors welcome bug reports of feature requests
+under active development, and authors welcome bug reports or feature requests
 via the GitHub repository.  The source for this manuscript is available on

### 2015-11-02 manuscript.Rnw
fix mini typo

--- 2015-11-02 manuscript.Rnw
@@
 where the development of \texttt{rotl} was started. We would also like to thank
-Scott Chamberlain (rOpenSci) for providing a thorough code review. FM was
+Scott Chamberlain (rOpenSci) for providing a thorough code review and Shinichi
+Nakagawa and Alistair Senior for their help in developing the package's
+meta-analysis vignette. DJW was supported by NIH Grant R01-GM101352. FM was
 supported by iDigBio, and therefore this material is based upon work supported

### 2016-03-08 manuscript.Rnw
add archive of rotl code used in manuscript
Start on revision suggested by AE

@@
 Phylogenetic information retrieved from OTL is converted into
-\texttt{ape::phylo} objects by \texttt{rotl} using the NEXUS Class Library
+\texttt{phylo} objects by \texttt{rotl} using the NEXUS Class Library
 \citep[NCL,][]{Lewis2003} as implemented in the \texttt{rncl} package
@@
 efficient parsing of large trees that may contain singleton nodes labeled with
-taxonomic information (i.e., monotypic taxon).
+taxonomic information (i.e. a monotypic taxon).
 
-The package is well-documented, and includes two package
+The package is well-documented, and includes three package
 vignettes (documents that demonstrate the use of the package and contain executable
 R code). There is also an extensive test-suite that covers both the internal
-functions that \texttt{rotl} uses to connect to OTL and public functions users
+functions that \texttt{rotl} uses to connect to OTL and public functions that users
 apply to access and process data.
@@
 
-To get the relationships among a set of taxa from the Open Tree, the taxa first
-need to be matched against the OTL taxonomy (OTT) using the TNRS. This step
-retrieves the identifiers that will be used to extract the relationships for the
-set of requested taxa.
-
-To illustrate how to obtain relationships from a set of taxa, we will rely on
-OTL to draw a phylogenetic tree for a set of model organisms. First, we use the function
-\texttt{tnrs\_match\_names} to match the taxon names to their Open Taxonomy identifiers.
+Before a researcher can use the Open Tree to retrieve relationships among a set
+of taxa, they first need to match the taxon names in their dataset with records
+in the Open Tree Taxonomy (OTT). The TNRS allows users to search for taxon
+names, and retreive identifiers for each matching taxon. We demonstrate the use
+of the TNRS within \texttt{rotl} by searching for taxonomic records associated
+with several model organisms.
 
@@
 and animals that are regulated by different nomenclatural codes (ICNP, ICN, and
-ICZN, respectively), both OTL and \texttt{rotl} provide tools to deal with potential
-hemihomonyms. The argument \texttt{context\_name} can be used to limit potential
+ICZN, respectively), both OTL and \texttt{rotl} provide tools to deal with names
+that may represent valid taxa in more than one code. The argument \texttt{context\_name} can be used to limit potential
 matches to a taxonomic group such as ``Animals'' (see the function
@@
 relationships among these taxa. In turn, the tree can be plotted directly as it
-is returned as an \texttt{ape::phylo} object (Figure~\ref{fig:plot_taxa}).
+is returned as \texttt{phylo} object (Figure~\ref{fig:plot_taxa}).
 
@@
 study. Alternatively, the function \texttt{get\_study} returns all the trees (by
-default as \texttt{ape::phylo} objects) associated with a particular
+default as \texttt{phylo}) associated with a particular
 study. Additional details about the study (i.e., its metadata) can be obtained
@@
 (\url{https://cran.r-project.org/package=rotl}), and development versions are
-available from GitHub (\url{https://github.com/ropensci/rotl}). The package is
-under active development, and authors welcome bug reports or feature requests
+available from GitHub (\url{https://github.com/ropensci/rotl}). This manuscript
+was built using version 0.4.1.999-rotl-ms.1
+(\url{https://github.com/ropensci/rotl/tree/v0.4.1.999-rotl-ms.1}). The package
+is under active development, and authors welcome bug reports or feature requests
 via the GitHub repository.  The source for this manuscript is available on

### 2016-03-08 reply_to_reviewers.md
Add reply document

Dear Professor Freckleton,

We thank the associate editor and reviewer for their comments, which have
improved our manuscript and the software it describes. Here provide detailed
replies to specific comments.

## Associate Editor

>The only code related thing that jumped out at me was to avoid
>`httr::content(req)` but to check the returned media type is as
>expected and do the json to R conversion manually (see
>`?httr::content`).

Thank you for this suggestion, we now only call httr::content to extract plain
text from request objects, and parse the results manually.

> l. 82: indicate that the api version can be selected?

> l. 97, and elsewhere: It's awkward in R but I don't think that
>`ape::phylo` is a good way of referring to the phylo class given it's
>not strictly namespaced.  "ape's `phylo` objects" perhaps?

In the first instance we now say "and are represented using the `ape` tree structure 
(objects of class `phylo`)". Elsewhere we simply refer to `phylo` objects.

>l. 100: the use of ncl is nice, especially with treatment of things
>like singletons.  Are these collapsed on import?  I believe in some
>versions of ape these will crash R when plotted.

[ francois? ]

>l. 102: "two vignettes" -> "three vignettes"
Done

>l. 109: here, and in a few places the paper gets very acronym heavy
>"OTL taxonomy (OTT) using the TNRS".  Consider rephrasing to get the
>intent across here (especially as I think in this is a really key
>point that may be obscured by all the acronyms) -- the data you have
>and the data that the service have need to be reconciled.

We have expanded this section a little, hopefully making the context clearer:

_Before a researcher can use the Open Tree to retrieve relationships among a set
of taxa, they first need to match the taxon names in their dataset with records
in the Open Tree Taxonomy (OTT). The TNRS allows users to search for taxon
names, and retreive identifiers for each matching taxon. We demonstrate the use
of the TNRS within `rotl` by searching for taxonomic records associated
with several model organisms._

>l. 118: "henihomonyms" may warrant a definition or replacement with
>something simpler.

replaced with "names that may represent valid taxa in more than one code."

>l. 136: it would be nice to know how many studies have been entered
>here, especially as a fraction of published trees.

[Ask Karen?]

>l. 143, code block: The "ot:focalCladeOTTTaxonName" bit here warrants
>some clarification, especially as to where a list of valid properties
>can be found.  This is not indicated in the help for
>`studies_find_studies` so a link there too (I presume to the OTL API
>docs) would be nice.

This is now explicitedly included in the documentation of
`studies_find_studies`. [djw note: do we want to reword this paragraph to make
it even clearer?]

>l. 164: Is there interaction here with taxize for TNRS?

Not at present, though there certainly could be.

### 2016-03-10 manuscript.Rnw
[fix #10] add numbers of trees and studies stored
emphasize that rncl collapses singletons
making the resulting compatible with APE

@@
 
-Phylogenetic information retrieved from OTL is converted into
-\texttt{phylo} objects by \texttt{rotl} using the NEXUS Class Library
-\citep[NCL,][]{Lewis2003} as implemented in the \texttt{rncl} package
+Phylogenetic information retrieved from OTL is converted into \texttt{phylo}
+objects by \texttt{rotl} using the NEXUS Class Library \citep[NCL,][]{Lewis2003}
+as implemented in the \texttt{rncl} package
 (\url{https://cran.r-project.org/package=rncl}). Using NCL provides robust and
 efficient parsing of large trees that may contain singleton nodes labeled with
-taxonomic information (i.e. a monotypic taxon).
+taxonomic information (i.e., a monotypic taxon). Singleton nodes are collapsed
+after the tree has been parsed, making the resulting \texttt{phylo} object
+compatible with all functions from the \texttt{ape} package.
 
@@
 been submitted through the curator interface, and identify the trees that
-contribute to the synthetic tree. These trees constitute a useful resource to
-reproduce or expand on a previously published analysis, or to explore how the
-elucidation of relationships within a clade has changed through time.
+contribute to the synthetic tree. As of March 2016, the Open Tree of Life
+project stores 7755 trees from 3399 studies (each having between 0 and 61
+trees), and 477 of these trees are used to assemble the synthetic tree. These
+trees constitute a useful resource to reproduce or expand on a previously
+published analysis, or to explore how the elucidation of relationships within
+a clade has changed through time.
 

### 2016-03-10 manuscript.Rnw
fix minor typo

@@
 adoption of open science practices in general. For instance, the
-rOpenSci-developed packge \texttt{TreeBase} \citep{Boettiger2012TB} allows users
+rOpenSci-developed package \texttt{TreeBase} \citep{Boettiger2012TB} allows users
 to access phylogenies stored in treeBASE (\url{www.treebase.org}).

### 2016-03-10 manuscript.Rnw
add paragraph about use of rotl for ecologists

@@
 \citep{Varela2015}), such that data associated with taxa in a given phylogeny
-can be obtained directly in R.  These packages, combined with the language's
-support for literate programming (\citealt{Knuth1984}; \citealt{Xie2014}), make
-R an ideal platform for reproducible research in phylogenetics and comparative
-biology, as they allow a complete record of the steps taken in gathering,
-processing, and analyzing a given data set to be produced.
+can be obtained directly in R.
+
+In ecology, the development of the field of community phylogenetics
+\citep{Webb2002} has created a need for researchers to have access to the
+evolutionary relationships of species making up communities. The relative
+contributions of the role of the environment (e.g., habitat filtering), and of
+biotic interactions (e.g., competitive exclusion), are inferred from the
+distribution of taxa on a phylogeny composed from species occurring at larger
+spatial scale. R packages and other software have been developed to generate
+phylogenies from species lists using taxonomic information or DNA
+sequences. These phylogenies can then be used for community phylogenetics
+analyses (e.g., \citep{Pearse2013}, \citep{Kembel}, \citep{Webb}), but as OTL
+becomes more comprehensive, it could become a valuable resource for ecologists.
+
+These packages, combined with the language's support for literate programming
+(\citealt{Knuth1984}; \citealt{Xie2014}), make R an ideal platform for
+reproducible research in phylogenetics and comparative biology, as they allow a
+complete record of the steps taken in gathering, processing, and analyzing a
+given data set to be produced.
 

### 2016-03-10 manuscript.Rnw
move citation at end of the sentence for clarity

@@
 current understanding of phylogenetic relationships across all organisms on
-Earth \citep{Hinchliff2015}, while providing tools and services that facilitate
-access to this information. OTL combines taxonomic information that serves as
-the backbone for the phylogenetic relationships, and published phylogenies to
-elucidate relationships among taxa. This combination of information is used to
-structure the comprehensive synthetic tree. Studies can be contributed to the
-synthetic tree through a curator interface
+Earth while providing tools and services that facilitate access to this
+information \citep{Hinchliff2015}. OTL combines taxonomic information that
+serves as the backbone for the phylogenetic relationships, and published
+phylogenies to elucidate relationships among taxa. This combination of
+information is used to structure the comprehensive synthetic tree. Studies can
+be contributed to the synthetic tree through a curator interface
 (\url{https://tree.opentreeoflife.org/curator}), allowing the synthetic tree to

### 2016-03-10 manuscript.Rnw
list versions of packages used to build figure 3

@@
 integration of phylogenetic data retrieved using \texttt{rotl} with other data
-here. We show how we can obtain a map of the occurrences for some of the cat
-species (genus \textit{Lynx}) included in the phylogeny above using records for
-these species found in GBIF (Figure~\ref{fig:cat_map}; code available in the
-Appendix).
+here.
 
 <<cat_map, purl=TRUE, echo=FALSE, size='small', fig.height=4.2, results='hide', message=FALSE, cache=TRUE, fig.cap='GBIF records for the species in \\textit{Lynx} included in the phylogeny associated with the study by \\citealt{Johnson2006}.'>>=
-## load required packages
-library(rotl)
-library(rgbif)
-library(ggplot2)
-library(maps)
+## load required packages. Versions used for the manuscript listed below.
+library(rotl)     # v0.5.0
+library(rgbif)    # v0.9.2
+library(ggplot2)  # v2.1.0
+library(maps)     # v3.1.0
 

### 2016-03-10 manuscript.Rnw
[fix #13] clarify diff betwwen vignette and appendix

@@
 
-The package vignettes provide two demonstrations of the integration of a
-phylogeny and data associated with the taxa it represents. Specifically, the ``Data
-mashups'' vignette provides an example of how data associated with tip-taxa can
-be gathered and visualized. Another vignette titled ``meta-analysis''
-demonstrates how a complete analysis, including the gathering of data and a
-phylogeny, can be performed in a single R session. We further demonstrate the
-integration of phylogenetic data retrieved using \texttt{rotl} with other data
-here.
+Here we show how we can obtain a map of the occurrences for a subset of the cat
+species (genus \textit{Lynx}) that were included in the phylogeny retrieved from
+the Felidae study above. We extract the species names from the phylogeny, and use
+them to ask for the records for these species found in GBIF
+(Figure~\ref{fig:cat_map}). We include the code to reproduce this figure in the
+Appendix.
+
+In addition of an introduction on how to use the package, \texttt{rotl} includes
+two vignettes that demonstrate how to integrate a phylogeny and data associated
+with the taxa it represents. Specifically, the ``Data mashups'' vignette
+provides an example of how to retrieve a phylogeny for species a researcher may
+have data for, and visualize both the phylogeny and data associated with the
+species at the tips. The other vignette titled ``Meta-analysis'' demonstrates how a complete
+comparative method analysis, including the gathering of data and a phylogeny, can be
+performed in a single R session. We reproduce a published meta-analysis testing
+for differential investment in male and female offsprings among 51 species of
+birds. As new versions of the OTL API and \texttt{rotl} are released,
+these vignettes will be kept up-to-date.
 

### 2016-03-10 manuscript.Rnw
[fix #11] add link to definition of properties
change wording slightly

@@
 available through the output of the function
-\texttt{studies\_properties}. Typically, users will want to search for studies
-or trees based on taxon names (or their OTT identifiers), but other
-criteria such as the title of the publication can be used. Here we demonstrate
-how to look for and retrieve trees for studies focusing on the family Felidae
-(Figure~\ref{fig:plot_cats}).
+\texttt{studies\_properties}. The meaning of these properties is described at:
+\url{https://github.com/OpenTreeOfLife/phylesystem-api/wiki/NexSON}. Typically,
+users will want to search for studies or trees based on taxon names (or their
+OTT identifiers), but other criteria such as the title of the publication can be
+used. Here we demonstrate how to look for and retrieve trees for studies
+focusing on the family Felidae (Figure~\ref{fig:plot_cats}).
 
@@
 Here we show how we can obtain a map of the occurrences for a subset of the cat
-species (genus \textit{Lynx}) that were included in the phylogeny retrieved from
-the Felidae study above. We extract the species names from the phylogeny, and use
-them to ask for the records for these species found in GBIF
-(Figure~\ref{fig:cat_map}). We include the code to reproduce this figure in the
+species that were included in the phylogeny retrieved from the Felidae study
+above (genus \textit{Lynx}). We extract the species names from the phylogeny,
+and use them to ask for the records for these species found in GBIF
+(Figure~\ref{fig:cat_map}). We include the code to reproduce this figure in
 Appendix.

### 2016-03-10 manuscript.Rnw
add that API versions can be selected
add mention of trait databases

@@
 \texttt{rotl} gives users access to all endpoints provided by version 2 of the
-OTL APIs. Phylogenetic trees served by the API can be imported directly into
+OTL APIs, and other versions of the API can be selected by the user as they
+become available. Phylogenetic trees served by the API can be imported directly into
 R's memory and are represented using the \texttt{ape} \citep{Paradis2004} tree
@@
 
+As trait databases are becoming increasingly more common, and interfaces to the
+data they contains are being developed as R packages \citep[e.g., the package
+\texttt{traits}][]{Chamberlain2015}, \texttt{rotl} provides a
+way to easily retrieve phylogenetic information for species trait data that are available.
+
 In addition of an introduction on how to use the package, \texttt{rotl} includes

### 2016-03-10 reply_to_reviewers.md
finish addressing AE's comments, and R1 comments

@@
 We thank the associate editor and reviewer for their comments, which have
-improved our manuscript and the software it describes. Here provide detailed
+improved our manuscript and the software it describes. Here we provide detailed
 replies to specific comments.
@@
 
-Thank you for this suggestion, we now only call httr::content to extract plain
+Thank you for this suggestion, we now only call `httr::content` to extract plain
 text from request objects, and parse the results manually.
@@
 > l. 82: indicate that the api version can be selected?
+
+We added: _and other versions of the API can be selected by the user as they
+become available._
 
@@
 
-[ francois? ]
+We added: _Singleton nodes are collapsed after the tree has been parsed, making
+the resulting \texttt{phylo} object compatible with all functions from the
+\texttt{ape} package._
 
 >l. 102: "two vignettes" -> "three vignettes"
+
 Done
@@
 
-[Ask Karen?]
+We added this information. See 1st paragraph of "Getting trees from studies".
 
@@
 This is now explicitedly included in the documentation of
-`studies_find_studies`. [djw note: do we want to reword this paragraph to make
-it even clearer?]
+`studies_find_studies`, and a link to the definition of the properties is
+provided in the manuscript.
 
@@
 
-Not at present, though there certainly could be.
+Not at present, but it's currently in progress. See:
+https://github.com/ropensci/taxize/issues/517
+
+## Reviewer 1
+
+> One thing I thought was missing was a discussion about the importance of rolt
+> to the field of ecology.
+
+We added a paragraph (l. xxx-xxx) that describes how `rotl` can be used as a
+data source for community phylogenetics.
+
+> 1) Can user generated trees be easily incorporated into an output tree from
+> rotl? Perhaps by implementing some of the algorithms used to integrate the
+> studies that comprise the OTL? For example, if I generated a tree for a genus
+> that was not yet in the OTL, could I add it into its family in the rotl
+> interface, or would I have to go through the curator interface?
+
+It's an interesting idea. If a researcher wants to include a better resolved
+tree than what is available in Open Tree, it is relatively straightforward to
+swap a monophyletic group by another one using the functions provided by
+`ape`. However, we feel that providing this kind of functionality falls beyond the
+scope of our package, whose main purpose is to provide an interface to the OTL
+API.
+
+> 2) I was a bit confused about why the authors chose to include the GBIF
+> example in the appendix, but not the other vignettes. Why not just add it as
+> another vignette, or include the other vignettes as appendices as well? A
+> small detail perhaps, but this makes the paragraph before the concluding
+> remarks a bit vague.
+
+We included the code for the GBIF example as an appendix as it seemed too long
+to be included in the text of the manuscript, and not long enough to warrant its
+own vignette. We didn't include the other vignettes in appendix to ensure that
+we can update the code they include in case the API changes.
+
+> 3) How might rolt be combined with trait databases? This was briefly mentioned
+> at the end of the vignette “Connecting data to Open Tree trees”, but might be
+> mentioned in the manuscript as well (for example, in the subsection “How does
+> rotl fit into the R package ecosystem?”).
+
+We added a paragraph to highlight that rotl can be used to provide a
+phylogenetic context to available trait data.
+
+> Page 1, paragraph 1: A comment about the importance of phylogenies for ecology
+> could be added here…
+
+Done
+
+> Page 2, paragraph 3, lines 39-41: This sentence is a bit awkward—perhaps
+> remove the comma?
+
+Done
+
+> Page 3, paragraph 1, line 58: Other recently developed R packages that relate
+> to rotl might be mentioned, including taxize (taxonomy congruence),
+> phyloGenerator (generate phylogenies for ecologists) and pez (for calculating
+> suites of metrics for community phylogenetics).
+
+We added information about related R packages that can be used for community
+ecology.
+
+> Page 4, point #2: So not all studies that can be queried through the API were
+> included in the OTL? Perhaps this can be explained a bit more here (or in the
+> “Getting trees from studies” subsection).
+
+We added the number of studies that have been submitted through the curator
+app, and the number of studies that contribute to the synthetic tree to clarify
+this point.
+
+> Page 9, line 150: What type of metadata can be extracted from a study? Is it
+> just the citation (which is the example given in “How to use rotl?” vignette)?
+
+There is a lot more information associated with each study. We re-worded the
+paragraph to mention some of them.
+
+> Page 10, paragraph 2, line 169: “other data” is a bit vague. Integration with
+> other data, for example species distribution data, is demonstrated in the
+> appendix (but not the main manuscript)
+
+We re-wrote this paragraph to be more specific about the type of data that can
+be linked to the phylogenies storred in `rotl`.

### 2016-03-10 manuscript.Rnw
expand on study metadata available

@@
 default as \texttt{phylo}) associated with a particular
-study. Additional details about the study (i.e., its metadata) can be obtained
+study. Metadata about the study (e.g., citation information, information about
+the curator for the study, and other technical information regarding the import
+of this study) can be obtained
 using the function \texttt{get\_study\_meta}.

### 2016-03-11 manuscript.Rnw
fix and add citations

@@
 spatial scale. R packages and other software have been developed to generate
-phylogenies from species lists using taxonomic information or DNA
-sequences. These phylogenies can then be used for community phylogenetics
-analyses (e.g., \citep{Pearse2013}, \citep{Kembel}, \citep{Webb}), but as OTL
-becomes more comprehensive, it could become a valuable resource for ecologists.
+phylogenies from species lists using taxonomic information or DNA sequences
+\citep[e.g.,][]{Webb2008,Pearse2013}. These phylogenies can then be used for
+community phylogenetics analyses \citep[e.g.,][]{Pearse2015},
+\citep{Kembel2010}, but they are often incomplete or not resolved enough. As OTL
+becomes more comprehensive, and its taxonomic resolution increases, it could
+become a valuable resource for ecologists seeking to use phylogenetic
+information in their research.
 
@@
 taxa matched are the correct ones. Here, there is no ambiguity in the taxa
-matched; however, as OTT includes taxa from bacteria, plants,
-and animals that are regulated by different nomenclatural codes (ICNP, ICN, and
-ICZN, respectively), both OTL and \texttt{rotl} provide tools to deal with names
-that may represent valid taxa in more than one code. The argument \texttt{context\_name} can be used to limit potential
-matches to a taxonomic group such as ``Animals'' (see the function
-\texttt{tnrs\_contexts} for a complete list of possible options). When this strategy
-cannot be used (as in the present example, where the tree encompasses
-multiple domains), the function \texttt{inspect} lists alternative matches for a taxon
-name, and \texttt{update} replaces it in the results. An example of this
-approach is provided in the vignette ``How to use \texttt{rotl}?'' that
-accompanies the package.
+matched; however, as OTT includes taxa from bacteria, plants, and animals that
+are regulated by different nomenclatural codes (ICNP, ICN, and ICZN,
+respectively), both OTL and \texttt{rotl} provide tools to deal with names that
+may represent valid taxa in more than one code. The argument
+\texttt{context\_name} can be used to limit potential matches to a taxonomic
+group such as ``Animals'' (see the function \texttt{tnrs\_contexts} for a
+complete list of possible options). When this strategy cannot be used (as in the
+present example, where the tree encompasses multiple domains), the function
+\texttt{inspect} lists alternative matches for a taxon name, and \texttt{update}
+replaces it in the results. An example of this approach is provided in the
+vignette ``How to use \texttt{rotl}?'' that accompanies the package.
 
 By default, approximate matching is enabled when attempting to match taxonomic
-names to their OTT identifiers. Additionally, taxonomic synonyms
-are included in OTT, allowing researchers to match correct
-identifiers for taxon names that might include misspellings or synonyms. These
-features will facilitate the tedious data cleaning process often needed when
-mapping taxon names. In the example provided, both \textit{Escherichia coli}
-and \textit{Saccharomyces cerevisiae} are misspelled, but the OTL TNRS finds the
-correct match for these taxa.
+names to their OTT identifiers. Additionally, taxonomic synonyms are included in
+OTT, allowing researchers to match correct identifiers for taxon names that
+might include misspellings or synonyms. These features will facilitate the
+tedious data cleaning process often needed when mapping taxon names. In the
+example provided, both \textit{Escherichia coli} and \textit{Saccharomyces
+  cerevisiae} are misspelled, but the OTL TNRS finds the correct match for these
+taxa.
 
@@
 cat_studies <- studies_find_studies(property = "ot:focalCladeOTTTaxonName",
-                                    value = "felidae")
-cat_studies
+                                    value = "Felidae", exact = TRUE)cat_studies
 @
@@
 data they contains are being developed as R packages \citep[e.g., the package
-\texttt{traits}][]{Chamberlain2015}, \texttt{rotl} provides a
-way to easily retrieve phylogenetic information for species trait data that are available.
+\texttt{traits}][]{Chamberlain2015}, \texttt{rotl} provides a way to easily
+retrieve phylogenetic information for species trait data that are available.
 
@@
 have data for, and visualize both the phylogeny and data associated with the
-species at the tips. The other vignette titled ``Meta-analysis'' demonstrates how a complete
-comparative method analysis, including the gathering of data and a phylogeny, can be
-performed in a single R session. We reproduce a published meta-analysis testing
-for differential investment in male and female offsprings among 51 species of
-birds. As new versions of the OTL API and \texttt{rotl} are released,
-these vignettes will be kept up-to-date.
+species at the tips. The other vignette titled ``Meta-analysis'' demonstrates
+how a complete comparative method analysis, including the gathering of data and
+a phylogeny, can be performed in a single R session. We reproduce a published
+meta-analysis testing for differential investment in male and female offsprings
+among 51 species of birds. As new versions of the OTL API and \texttt{rotl} are
+released, these vignettes will be kept up-to-date.
 

### 2016-03-11 manuscript.Rnw
caught missing newline

@@
 cat_studies <- studies_find_studies(property = "ot:focalCladeOTTTaxonName",
-                                    value = "Felidae", exact = TRUE)cat_studies
+                                    value = "Felidae", exact = TRUE)
+cat_studies
 @

### 2016-03-11 manuscript.Rnw
Update cat_tree text to reflect new trees in OTL :joy:

@@
 
-Currently only one study focused on this family is available from OTL, and a
-single tree is associated with it. We can then retrieve the study and tree
-identifiers, and pass them to the function \texttt{get\_study\_tree} to have the
-tree in memory:
+Currently there are seven studies focused on this family available from OTL, and a
+single tree is associated with each one. We can then retrieve the study and tree
+identifiers associated with the first of these studies, and pass them to the 
+function \texttt{get\_study\_tree} to have the tree in memory:
 

### 2016-03-21 install_fonts.R
attempt at using Travis

@@ -1,0 +1,5 @@
+install_fonts <- function() {
+    dir.create("~/.fonts")
+    system("git clone git@github.com:fmichonneau/latex-fonts.git ~/.fonts")
+    system("fc-cache -f -v")
+}

### 2016-03-21 manuscript.Rnw
incorporate Ross's comments on pre-print

--- 2016-03-11 manuscript.Rnw
@@
 phylogenetic data makes it difficult for researchers to access, synthesize, and
-integrate this information into their own research (\citealt{Stoltzfus2012};
-\citealt{Drew2013}; but see \citealt{Cranston2014} for suggestions of best
+integrate this information into their own research (\citealt{Stoltzfus2012,
+  Drew2013, Magee2014}; but see \citealt{Cranston2014} for suggestions of best
 practices).
@@
 These packages, combined with the language's support for literate programming
-(\citealt{Knuth1984}; \citealt{Xie2014}), make R an ideal platform for
+(\citealt{Knuth1984}; \citealt{Xie2014}), make R a comprehensive platform for
 reproducible research in phylogenetics and comparative biology, as they allow a

### 2016-03-21 manuscript.Rnw
fix typos
include Scott's comments on pre-print

@@
 users to query and access data from the web (e.g. \texttt{rFISHBASE}
-\citep{Boettiger2012}, \texttt{rAvis} \citep{Varela2014}, and \texttt{paleoDB}
+\citep{Boettiger2012}, \texttt{rAvis} \citep{Varela2014}, and \texttt{paleobioDB}
 \citep{Varela2015}), such that data associated with taxa in a given phylogeny
@@
 
-Here we present \texttt{rotl}, and R package that allows users to download
+Here we present \texttt{rotl}, an R package that allows users to download
 phylogenetic and taxonomic data from the OTL directly in R. \texttt{rotl} takes
@@
 of taxa, they first need to match the taxon names in their dataset with records
-in the Open Tree Taxonomy (OTT). The TNRS allows users to search for taxon
-names, and retreive identifiers for each matching taxon. We demonstrate the use
-of the TNRS within \texttt{rotl} by searching for taxonomic records associated
-with several model organisms.
+in the Open Tree Taxonomy (OTT). OTL's taxonomic names resolution service (TNRS)
+combines information from multiple services (e.g., National Center for
+Biotechnology Information [NCBI], World Register for Marine Species [WoRMS],
+Global Biodiversity Information Facility [GBIF]) and allows users to search for
+taxon names, and retreive identifiers for each matching taxon. We demonstrate
+the use of the TNRS within \texttt{rotl} by searching for taxonomic records
+associated with several model organisms.
 

### 2016-03-21 manuscript.Rnw
acknowledge Scott & Ross for commenting pre-print

@@
 where the development of \texttt{rotl} was started. We would also like to thank
-Scott Chamberlain (rOpenSci) for providing a thorough code review and Shinichi
-Nakagawa and Alistair Senior for their help in developing the package's
-meta-analysis vignette. DJW was supported by NIH Grant R01-GM101352. FM was
-supported by iDigBio, and therefore this material is based upon work supported
-by the National Science Foundation’s Advancing Digitization of Biodiversity
-Collections Program (Cooperative Agreement EF-1115210).
+Scott Chamberlain (rOpenSci) for providing a thorough code review, Scott
+Chamberlain and Ross Mounce for commenting on the pre-print version of this
+manuscript, and Shinichi Nakagawa and Alistair Senior for their help in
+developing the package's meta-analysis vignette. DJW was supported by NIH Grant
+R01-GM101352. FM was supported by iDigBio, and therefore this material is based
+upon work supported by the National Science Foundation’s Advancing Digitization
+of Biodiversity Collections Program (Cooperative Agreement EF-1115210).
 

### 2016-03-21 manuscript.Rnw
Include reference to phangorn as core R phylo pkg

@@
 analysis. The R packages \texttt{ape} \citep{Paradis2004}, \texttt{phylobase}
-\citep{phylobase-0.8.0}, and \texttt{RNeXML} \citep{Boettiger2015} each provide
+\citep{phylobase-0.8.0}, \texttt{phangorn} \citep{schliep2011} and \texttt{RNeXML} \citep{Boettiger2015} each provide
 functions to import and manipulate trees within R and save the results in
@@
 In recent years, R has become an essential part of the toolbox of many
-researchers in evolutionary biology. R greatly facilitates the analysis of large
-datasets, and allows researchers to combine methods in novel ways because many
-methods for comparative analyses are implemented, and because it is a relatively
-easy to use programming language. Additionally, as more data are made available
-online and accessible using APIs, several packages have been developed to
-interact and download these datasets directly in R, thereby enabling direct and
-reproducible analyses.  Notably, the organization rOpenSci
+researchers in evolutionary biology and ecology. R greatly facilitates the 
+analysis of large datasets, and allows researchers to combine methods in novel
+ways because many methods for comparative analyses are implemented, and because
+it is a relatively easy to use programming language. Additionally, as more data 
+are made available online and accessible using APIs, several packages have been 
+developed to interact and download these datasets directly in R, thereby 
+enabling direct and reproducible analyses.  Notably, the organization rOpenSci
 (\url{https://ropensci.org}) has fostered a community of researchers who develop

### 2016-03-21 manuscript.Rnw
update 'traits' citation

--- 2016-03-21 manuscript.Rnw
@@
 data they contains are being developed as R packages \citep[e.g., the package
-\texttt{traits}][]{Chamberlain2015}, \texttt{rotl} provides a way to easily
+\texttt{traits}][]{Chamberlain2016}, \texttt{rotl} provides a way to easily
 retrieve phylogenetic information for species trait data that are available.

### 2016-03-21 manuscript.Rnw
add dev version number in code appendix

--- 2016-03-21 manuscript.Rnw
@@
 ## load required packages. Versions used for the manuscript listed below.
-library(rotl)     # v0.5.0
+library(rotl)     # v0.5.0.901
 library(rgbif)    # v0.9.2

### 2016-03-21 manuscript.Rnw
[fix #15] removing paragraph about gol endpoints

@@
 \end{enumerate}
-
-Additionally, the project provides access to other data, the Graph of Life, that
-stores information about the underlying tree alignment graph \citep{Smith2013}
-used to synthesize the various sources of information from which the Open Tree
-is assembled. Data from this API are not intended for normal users, and its
-use will not be demonstrated in this paper, although \texttt{rotl} has functions to
-retrieve data it provides.
 

### 2016-03-21 install_fonts.R
use https instead of ssh to clone repo

@@ -1,5 +1,5 @@
 install_fonts <- function() {
     dir.create("~/.fonts")
-    system("git clone git@github.com:fmichonneau/latex-fonts.git ~/.fonts")
+    system("git clone https://github.com/fmichonneau/latex-fonts.git ~/.fonts")
     system("fc-cache -f -v")
 }

### 2016-03-21 manuscript.Rnw
add paragraphs about gol and collection endpoints

@@
 
-\texttt{rotl} gives users access to all endpoints provided by version 2 of the
-OTL APIs, and other versions of the API can be selected by the user as they
-become available. Phylogenetic trees served by the API can be imported directly into
-R's memory and are represented using the \texttt{ape} \citep{Paradis2004} tree
+\texttt{rotl} gives users access to the endpoints provided by version 2 of the
+APIs, and other versions of the APIs can be selected by the user as they become
+available (version 3 is scheduled to be released at the end of March 2016). Two
+other end-points are available in version 2 of the API: \emph{graph of life} and
+\emph{collections}. Data from the \emph{graph of life} endpoint are not intended
+for typical users, and will be removed in future releases. The \emph{collection}
+endpoint is not fully functional, and no functions are available in the current
+release of \texttt{rotl}. Future versions will make them available when the API
+implements them.
+
+Phylogenetic trees served by the API can be imported directly into R's
+memory and are represented using the \texttt{ape} \citep{Paradis2004} tree
 structure (objects of class \texttt{phylo}), or can be written to files in the

### 2016-03-21 manuscript.Rnw
fix typo, reflow paragraphs

@@
 
-The full Open Tree currently does not have any branch lengths associated with
+Currently, the synthetic tree does not have any branch lengths associated with
 it, therefore parametric comparative methods cannot be used directly on the
-subtrees returned by OTL (although the OTL treestore contains the raw
-published source phylogenies, complete with branch lengths and node
-annotations; see below). However, resources and methods are being developed to add
-branch lengths to these topological subtrees \citep[e.g.,][]{Ksepka2015} or use
-topological trees to identify phylogenetically equivalent species to increase
-overlap between chronograms and species trait data \citep{Pennell2015}. Without
-branch lengths, these subtrees are nonetheless useful to illustrate relationships
-among species, or to map traits on a phylogeny.
+subtrees returned by OTL (although the OTL treestore contains the raw published
+source phylogenies, complete with branch lengths and node annotations; see
+below). However, resources and methods are being developed to add branch lengths
+to these topological subtrees \citep[e.g.,][]{Ksepka2015} or use topological
+trees to identify phylogenetically equivalent species to increase overlap
+between chronograms and species trait data \citep{Pennell2015}. Without branch
+lengths, these subtrees are nonetheless useful to illustrate relationships among
+species, or to map traits on a phylogeny.
 
@@
 
-The package is well-documented, and includes three package
-vignettes (documents that demonstrate the use of the package and contain executable
-R code). There is also an extensive test-suite that covers both the internal
-functions that \texttt{rotl} uses to connect to OTL and public functions that users
-apply to access and process data.
+The package is well-documented, and includes three package vignettes (documents
+that demonstrate the use of the package and contain executable R code). There is
+also an extensive test-suite that covers both the internal functions that
+\texttt{rotl} uses to connect to OTL, and public functions that users apply to
+access and process data.
 
@@
 study. Alternatively, the function \texttt{get\_study} returns all the trees (by
-default as \texttt{phylo}) associated with a particular
-study. Metadata about the study (e.g., citation information, information about
-the curator for the study, and other technical information regarding the import
-of this study) can be obtained
-using the function \texttt{get\_study\_meta}.
+default as \texttt{phylo} objects) associated with a particular study. Metadata
+about the study (e.g., citation information, information about the curator for
+the study, and other technical information regarding the import of this study)
+can be obtained using the function \texttt{get\_study\_meta}.
 
@@
 tools and methods to facilitate the use of open data as well broaden the
-adoption of open science practices in general. For instance, the
-rOpenSci-developed package \texttt{TreeBase} \citep{Boettiger2012TB} allows users
-to access phylogenies stored in treeBASE (\url{www.treebase.org}).
-\texttt{rotl} contributes to this initiative, and greatly extends the number of
-taxa for which phylogenetic data can be retrieved within R, while allowing the
-data from OTL to be combined with other sources easily.
+adoption of open science practices in general \citep{Boettiger2015a}. For
+instance, the rOpenSci-developed package \texttt{TreeBase}
+\citep{Boettiger2012TB} allows users to access phylogenies stored in treeBASE
+(\url{www.treebase.org}).  \texttt{rotl} contributes to this initiative, and
+greatly extends the number of taxa for which phylogenetic data can be retrieved
+within R, while allowing the data from OTL to be combined with other sources
+easily.
 

### 2016-03-22 reply_to_reviewers.md

@@
 
-We added information about related R packages that can be used for community
-ecology.
+We added a paragraph about the use of phylogenetics in community ecology,
+inclding a description of R packages that can be used for these analysis [Pg
+Line]
 
@@
 
+"As of March 2016, the Open Tree of Life project stores 7755 trees from 3399 
+studies (each having between 0 and 61 trees), and 477 of these trees are used 
+to assemble the synthetic tree."
+
 > Page 9, line 150: What type of metadata can be extracted from a study? Is it
@@
 paragraph to mention some of them.
+
+"Metadata about the study (e.g., citation information, information about
+the curator for the study, and other technical information regarding the import
+of this study) can be obtained using the function `get_study_meta`
 

### 2016-03-22 manuscript.Rnw
update tag version used to generate the manuscript

@@
 The recognition of the importance of phylogenies to account for the statistical
-non-independence of species in comparative methods, and the recent development
-of methods to explore trait evolution or changes in diversification rates, have
-driven the need for accurate phylogenies. However, there is often a discrepancy
-between taxa targeted by comparative methods, and taxa for which phylogenies are
-available. We believe that by providing an easy-to-use interface to obtain
-phylogenies for an arbitrary set of taxa directly in R, \texttt{rotl} will be
-useful in a wide variety of contexts.
+non-independence of species in comparative methods, the recent development of
+methods to explore trait evolution or changes in diversification rates, and
+attempts to incorporate the evolutionary history of species forming ecological
+communities, have driven the need for accurate phylogenies. However, there is
+often a discrepancy between taxa targeted by studies wanting to use phylogenetic
+information, and taxa for which phylogenies are available. Typically, the latter
+result from focused studies of taxonomic groups, while the former encompass
+species found in a given geographical location or ecosystem. We believe that by
+providing an easy-to-use interface to obtain phylogenies for an arbitrary set of
+taxa directly in R, \texttt{rotl} will be useful in a wide variety of contexts.
 
@@
 available from GitHub (\url{https://github.com/ropensci/rotl}). This manuscript
-was built using version 0.4.1.999-rotl-ms.1
-(\url{https://github.com/ropensci/rotl/tree/v0.4.1.999-rotl-ms.1}). The package
+was built using version 0.5.0.901-rotl-ms.2
+(\url{https://github.com/ropensci/rotl/tree/v0.5.0.901-rotl-ms.2}). The package
 is under active development, and authors welcome bug reports or feature requests

### 2016-03-22 reply_to_reviewers.md
update letter to reviewers

--- 2016-03-10 reply_to_reviewers.md
@@
 We thank the associate editor and reviewer for their comments, which have
-improved our manuscript and the software it describes. Here we provide detailed
+improved our manuscript and the software it describes. Below we provide detailed
 replies to specific comments.
+
+The developers working on the Open Tree of Life APIs have informed us that a new
+version of the APIs is going to be released in the next few weeks. We don't
+expect these changes to impact what we describe in the manuscript, but small
+tweaks to the package might be needed.
+
+We look forward to hearing back from you.
+
+  François Michonneau on behalf of Joseph Brown and David Winter
 
@@
 
-replaced with "names that may represent valid taxa in more than one code."
+replaced with "_names that may represent valid taxa in more than one code._"
 
@@
 
-We added a paragraph (l. xxx-xxx) that describes how `rotl` can be used as a
-data source for community phylogenetics.
+We added a paragraph (l. 60-70) that describes how `rotl` can be used as a
+data source for community phylogenetics. We also mention community ecology in
+the concluding remarks (l. 202-210).
 
@@
 to be included in the text of the manuscript, and not long enough to warrant its
-own vignette. We didn't include the other vignettes in appendix to ensure that
-we can update the code they include in case the API changes.
+own vignette. We didn't include the other vignettes as appendices to ensure that
+we can update the code they include when the API changes.
 
@@
 
-We added a paragraph to highlight that rotl can be used to provide a
-phylogenetic context to available trait data.
+We added a paragraph to highlight that `rotl` can be used to provide a
+phylogenetic context to available trait data (l. 189-191).
 
@@
 
-We added information about related R packages that can be used for community
-ecology.
+We added a paragraph about the use of phylogenetics in community ecology,
+including a description of R packages that can be used for these analysis
+(l. 65-70).
 
@@
 
-We added the number of studies that have been submitted through the curator
-app, and the number of studies that contribute to the synthetic tree to clarify
-this point.
+We added the number of studies that have been submitted through the curator app,
+and the number of studies that contribute to the synthetic tree to clarify this
+point: "_As of March 2016, the Open Tree of Life project stores 7755 trees from
+3399 studies (each having between 0 and 61 trees), and 477 of these trees are
+used to assemble the synthetic tree._"
 
@@
 There is a lot more information associated with each study. We re-worded the
-paragraph to mention some of them.
+paragraph to mention some of them: "_Metadata about the study (e.g., citation
+information, information about the curator for the study, and other technical
+information regarding the import of this study) can be obtained using the
+function `get_study_meta`_."
 
@@
 We re-wrote this paragraph to be more specific about the type of data that can
-be linked to the phylogenies storred in `rotl`.
+be linked to the phylogenies storred in `rotl` (l. 168-169).

### 2016-03-22 manuscript.Rnw
fix line breaks

@@
 analysis. The R packages \texttt{ape} \citep{Paradis2004}, \texttt{phylobase}
-\citep{phylobase-0.8.0}, \texttt{phangorn} \citep{schliep2011} and \texttt{RNeXML} \citep{Boettiger2015} each provide
-functions to import and manipulate trees within R and save the results in
-standard data formats. Additional packages including \texttt{phytools}
-\citep{Revell2012}, \texttt{geiger} \citep{Pennell2014-geiger}, and
-\texttt{ggtree} \citep{Yu-ggtree}, allow users to analyze and visualize data in
-a phylogenetic context (see
-\url{https://cran.r-project.org/web/views/Phylogenetics.html} for a
+\citep{phylobase-0.8.0}, \texttt{phangorn} \citep{schliep2011} and
+\texttt{RNeXML} \citep{Boettiger2015} each provide functions to import and
+manipulate trees within R and save the results in standard data
+formats. Additional packages including \texttt{phytools} \citep{Revell2012},
+\texttt{geiger} \citep{Pennell2014-geiger}, and \texttt{ggtree}
+\citep{Yu-ggtree}, allow users to analyze and visualize data in a phylogenetic
+context (see \url{https://cran.r-project.org/web/views/Phylogenetics.html} for a
 comprehensive list of phylogenetics packages in R). In addition to packages for
@@
 users to query and access data from the web (e.g. \texttt{rFISHBASE}
-\citep{Boettiger2012}, \texttt{rAvis} \citep{Varela2014}, and \texttt{paleobioDB}
-\citep{Varela2015}), such that data associated with taxa in a given phylogeny
-can be obtained directly in R.
+\citep{Boettiger2012}, \texttt{rAvis} \citep{Varela2014}, and
+\texttt{paleobioDB} \citep{Varela2015}), such that data associated with taxa in
+a given phylogeny can be obtained directly in R.
 
@@
 \citep[e.g.,][]{Webb2008,Pearse2013}. These phylogenies can then be used for
-community phylogenetics analyses \citep[e.g.,][]{Pearse2015},
-\citep{Kembel2010}, but they are often incomplete or not resolved enough. As OTL
-becomes more comprehensive, and its taxonomic resolution increases, it could
-become a valuable resource for ecologists seeking to use phylogenetic
-information in their research.
+community phylogenetics analyses \citep[e.g.,][]{Pearse2015, Kembel2010}, but
+they are often incomplete or not resolved enough. As OTL becomes more
+comprehensive, and its taxonomic resolution increases, it could become a
+valuable resource for ecologists seeking to use phylogenetic information in
+their research.
 

### 2016-03-30 manuscript.Rnw
reformat abstract according to guidelines

@@
 
-While phylogenies have been getting easier to build, it has been difficult to
-re-use, combine, and synthesize the information they provide because published
-trees are often only available as image files, and taxonomic information is not
-standardized across studies. The Open Tree of Life (OTL) project addresses these
-issues by providing a digital tree that encompasses all organisms, built by
-combining taxonomic information and published phylogenies. The project also
-provides tools and services to query and download parts of this synthetic tree,
-as well as the source data used to build it. Here, we present \texttt{rotl}, an
-R package to search and download data from the Open Tree of Life directly in
-R. \texttt{rotl} uses common data structures allowing researchers to take
+\begin{enumerate}
+\item While phylogenies have been getting easier to build, it has been difficult
+  to re-use, combine, and synthesize the information they provide because
+  published trees are often only available as image files, and taxonomic
+  information is not standardized across studies.
+\item The Open Tree of Life (OTL) project addresses these issues by providing a
+  digital tree that encompasses all organisms, built by combining taxonomic
+  information and published phylogenies. The project also provides tools and
+  services to query and download parts of this synthetic tree, as well as the
+  source data used to build it. Here, we present \texttt{rotl}, an R package to
+  search and download data from the Open Tree of Life directly in R.
+\item \texttt{rotl} uses common data structures allowing researchers to take
   advantage of the rich set of tools and methods that are available in R to
-manipulate, analyze, and visualize phylogenies.
+  manipulate, analyze, and visualize phylogenies. Here, and in the vignettes
+  accompanying the package, we demonstrate how \texttt{rotl} can be used with
+  other R packages to analyze biodiversity data.
+\item As phylogenies are being used in a growing number of applications,
+  \texttt{rotl} facilitates access to phylogenetic data, and allows their
+  integration with statistical methods and data sources available in R.
+\end{enumerate}
 

### 2016-03-30 manuscript.Rnw
add AE and reviewer in acknowledgments

@@
 Chamberlain and Ross Mounce for commenting on the pre-print version of this
-manuscript, and Shinichi Nakagawa and Alistair Senior for their help in
-developing the package's meta-analysis vignette. DJW was supported by NIH Grant
-R01-GM101352. FM was supported by iDigBio, and therefore this material is based
-upon work supported by the National Science Foundation’s Advancing Digitization
-of Biodiversity Collections Program (Cooperative Agreement EF-1115210).
+manuscript, Shinichi Nakagawa and Alistair Senior for their help in developing
+the package's meta-analysis vignette, and Rich FitzJohn (Associate Editor) and
+an anonymous reviewer for comments on the manuscript. DJW was supported by NIH
+Grant R01-GM101352. FM was supported by iDigBio, and therefore this material is
+based upon work supported by the National Science Foundation’s Advancing
+Digitization of Biodiversity Collections Program (Cooperative Agreement
+EF-1115210).
 

### 2016-04-01 manuscript.Rnw
mention 1923 Fund in acknowledgments

@@
 an anonymous reviewer for comments on the manuscript. DJW was supported by NIH
-Grant R01-GM101352. FM was supported by iDigBio, and therefore this material is
-based upon work supported by the National Science Foundation’s Advancing
-Digitization of Biodiversity Collections Program (Cooperative Agreement
-EF-1115210).
+Grant R01-GM101352. FM was supported by 1923 Fund and by iDigBio, and therefore
+this material is based upon work supported by the National Science Foundation’s
+Advancing Digitization of Biodiversity Collections Program (Cooperative
+Agreement EF-1115210).
 

### 2016-04-20 manuscript.Rnw
update manuscript to latest rotl version

@@
 analysis. The R packages \texttt{ape} \citep{Paradis2004}, \texttt{phylobase}
-\citep{phylobase-0.8.0}, \texttt{phangorn} \citep{schliep2011} and
+\citep{phylobase-0.8.0}, \texttt{phangorn} \citep{schliep2011}, and
 \texttt{RNeXML} \citep{Boettiger2015} each provide functions to import and
@@
 
-\texttt{rotl} gives users access to the endpoints provided by version 2 of the
+\texttt{rotl} gives users access to the endpoints provided by version 3 of the
 APIs, and other versions of the APIs can be selected by the user as they become
-available (version 3 is scheduled to be released at the end of March 2016). Two
-other end-points are available in version 2 of the API: \emph{graph of life} and
-\emph{collections}. Data from the \emph{graph of life} endpoint are not intended
-for typical users, and will be removed in future releases. The \emph{collection}
-endpoint is not fully functional, and no functions are available in the current
-release of \texttt{rotl}. Future versions will make them available when the API
-implements them.
+available.
 

### 2016-04-26 manuscript.Rnw
update manuscript following rotl 3.0.0 release

@@
 <<plot_taxa, fig.height=4, cache=TRUE, purl=FALSE, fig.cap="The phylogenetic tree returned by OTL for the list of model species used as an example.">>=
-tree <- tol_induced_subtree(ott_ids = taxa[["ott_id"]])
+tree <- tol_induced_subtree(ott_ids = ott_id(taxa))
 plot(tree, cex = .8, label.offset = .1, no.margin = TRUE)
@@
 the study, and other technical information regarding the import of this study)
-can be obtained using the function \texttt{get\_study\_meta}.
+can be obtained using the functions \texttt{get\_study\_meta} and
+\texttt{study\_external\_IDs}.
 
@@
 it is a relatively easy to use programming language. Additionally, as more data
-are made available online and accessible using APIs, several packages have been
-developed to interact and download these datasets directly in R, thereby
+are made available online and accessible using web APIs, several packages have
+been developed to interact and download these datasets directly in R, thereby
 enabling direct and reproducible analyses.  Notably, the organization rOpenSci
@@
 instance, the rOpenSci-developed package \texttt{TreeBase}
-\citep{Boettiger2012TB} allows users to access phylogenies stored in treeBASE
-(\url{www.treebase.org}).  \texttt{rotl} contributes to this initiative, and
+\citep{Boettiger2012TB} allows users to access phylogenies stored in TreeBASE
+(\url{https://treebase.org}).  \texttt{rotl} contributes to this initiative, and
 greatly extends the number of taxa for which phylogenetic data can be retrieved
@@
 ## load required packages. Versions used for the manuscript listed below.
-library(rotl)     # v0.5.0.901
-library(rgbif)    # v0.9.2
+library(rotl)     # v3.0.0
+library(rgbif)    # v0.9.3
 library(ggplot2)  # v2.1.0
@@
 available from GitHub (\url{https://github.com/ropensci/rotl}). This manuscript
-was built using version 0.5.0.901-rotl-ms.2
-(\url{https://github.com/ropensci/rotl/tree/v0.5.0.901-rotl-ms.2}). The package
+was built using \texttt{rotl} 3.0.0
+(\url{https://github.com/ropensci/rotl/tree/v3.0.0}). The package
 is under active development, and authors welcome bug reports or feature requests

### 2016-07-19 create_appendix.R
final edits used at submission

@@ -2,7 +2,9 @@ knitr::purl("manuscript.Rnw", documentation = 0L)
 
 header <- c("---",
             "title: Appendix",
-            "output: pdf_document",
+            "output: ",
+            "  pdf_document:",
+            "    keep_tex: yes",
             "---",
             "\n",
             "Source code used to generate Figure 3.",
@@ -15,4 +17,4 @@ code_chunk <- readLines("manuscript.R")
 cat(header, code_chunk, footer, sep = "\n",
     file = "manuscript-appendix.Rmd")
 
-rmarkdown::render("manuscript-appendix.Rmd")
+rmarkdown::render("manuscript-appendix.Rmd", clean = FALSE)
